### 2025-02-10 paper/background.tex
Adding paper sources from the last year.

\section{Background\label{sec:background}}

Automatic differentiation has been around for many decades~\cite{early-ad1, early-ad2},
so it is well-understood at a conceptual level.  However,
a number of questions related to bringing AD into the context of
programming languages remain open.  Recent successes in machine learning
have spurred further interest in AD which has led to several new developments.
For the context of this paper, we focus on recent work that contributes to 
the perspective of balancing correctness guarantees and performance.
Our selection here is by no means exhaustive, for
a broader scope we refer the reader to~\cite{autodiff-survey}.

There has been a number of programming-language-oriented approaches that explain
how to add AD to a programming language of choice. Examples of these include
Futhark~\cite{futhark/sc22ad}, Haskell~\cite{ad-haskell}, and
Jax~\cite{ad-jax}. Furthermore, a number of machine learning
frameworks that incorporate AD have been proposed in recent years: TensorFlow~\cite{ad-tf},
PyTorch~\cite{ad-pytorch}, MXNet~\cite{ad-mxnet} and many more.
While in particular the dedicated frameworks tend to find widespread 
acceptance by practitioners, both, correctness and performance leave
two open questions: (i) how is it possible to
ensure that the AD algorithm is implemented correctly? (ii) if the
language or the framework do not perform as expected, what are the
options to solve this?  Unfortunately, for many cases the answer to
both questions is unsatisfying.  Most of the languages/frameworks do not
come with formal correctness guarantees, so one has to trust the
implementers of these tools.  One can run tests as well to gain trust 
in the implementation but that is far from a 
formal guarantee.  If one relies
on the AD provided by a chosen language/framework, and the generated code does not
perform well, one has to modify the language/framework, as these solutions
are tightly integrated with the tools. The problem here is that most of of these tools
have very large and sophisticated implementations typically comprising
of hundreds of thousands of lines of code.  Furthermore, these systems
often rely on sophisticated 
combinations of sub-systems that need to be fine-tuned to the executing hardware.

Another line of work studies high-level principles of AD using
category theory~\cite{ad-theor1, ad-theor2, ad-theor3}.
While this indeed comes with great correctness guarantees due to
some naturality principles, it is not always clear how to implement
this in a way that leads to efficiently executable specifications.  Also, the
entire treatment of index-safe tensors is unclear.

In~\cite{ad-elliott} the author proposes to view AD problem using
the language of cartesian categories.  It has been shown that
this approach can be used in practice by implementing the proposed
technique in Haskell.  Type classes are a vehicle to restrict expressions
that are differentiable.  There is a Haskell plugin that translates
expressions that are instances of the mentioned type classes into
categorical primitives, AD is performed on these and the code is reflected
back to Haskell.  This is a nice approach that makes it easy
to verify the correctness of the algorithm.  However, the treatment
of tensors and general extractability remains a little unclear.
While it is briefly mentioned that representable functors
are supported, it is unclear whether this is sufficient to
represent rank-polymorphic arrays with strict bound checks.
Also, correctness guarantees are inevitably restricted by the
Haskell type system, so we are likely to find invariants that
are inexpressible in that setup.

### 2025-02-10 paper/conclusions.tex
Adding paper sources from the last year.

\section{Conclusions\label{sec:conclusions}}

The paper demonstrates a technique of developing high performance applications with
strong correctness guarantees.  The key insight lies in using a proof assistant
in cooperation with a high-performance language of choice.  This gives a clear
separation of concerns that is very difficult to achieve within a single
language.  The proof assistant is used to design a specification, prove
all the correctness invariants of interest and performs an extraction
into a high-performance language of choice.  This may take a non-trivial
effort, but correctness \emph{is} demanding!

Having a trusted specification as well as entire code-generation pipeline
within a single dependently-typed framework is incredibly powerful.
As we have demonstrated at the example of the neural network, we can introduce domain-specific
optimisations and even swap the backend in case its performance is unsatisfying.
For our example, the entire framework that includes array theory, DSL,
optimisations and extraction is about 2000 lines of Agda code.  We managed
to introduce a new backend in about two days and match performance of
the hand-written code.

A lot of pieces that we have developed in this
paper can be reused in other numerical applications.  We used dependent
types to guarantee the absence of out-of-bound indexing, certain function being
inverses as well as well-scopedness and well-typedness of our DSL.
However, there are many more opportunities that we did not explore.
For example, one can prove the correctness of optimisations, relating
evaluation of optimised an non-optimised expressions.  We can provide
more guarantees when we run extraction, \eg{} we can formalise some
aspects of the backend language and relate them to our DSL.
As for the DSL itself, we can try extending it with internal let
constructions which should improve our C code generation as well as
facilitate optimisations of the derivatives.

There are indeed plenty of opportunities, but the key point is this.
Correctness and performance are competing requirements when it
comes to application design.  Therefore, such a cooperation between
correctness-oriented and performance-oriented tools is likely to
persist.  With this work we demonstrate that such cooperation
is feasible in practice.

### 2025-02-10 paper/intro.tex
Adding paper sources from the last year.

\section{Introduction\label{sec:intro}}

The year is 2024, and we still have to make a formidable choice between
correctness and performance for all the programming projects that we start.
Low level programming languages such as C or Fortran make it possible to leverage
intricate hardware features for the price of poor analysability and
correctness guarantees.  Dependently-typed systems such as Lean or Agda make it possible
to describe arbitrary invariants within the given program, yet they can
rarely generate high-performance code.

The idealist inside of us is exasperated, because there should be a perfect
solution that caters for both cases.  However, the practitioner within us
proposes a different perspective.
Instead of using a single language, we may use two languages
in cooperation.  Specifically, we envision an alliance between a
proof assistant and some high-performance language of our choice.

To explore this idea, we investigate the concrete problem of automatic differentiation (AD) which
is often found in machine learning applications.  This is a convenient case study
as it comes with the following challenges.  From the correctness perspective,
it is crucially important to track the shapes and ranks of the tensors,
guaranteeing the absence of out-of-bound indexing.
This is a very common source of errors that can be incredibly difficult
to find.   Secondly, we have to compute derivatives of the given tensor expressions,
preserve safe indexing guarantees while we do so, and we have to be able to translate
the computed expressions into some high-performance language.
As machine learning applications
are known to be numerically intensive problems, our performance challenge lies
in running the program as fast as we can on the chosen hardware architecture.

We follow~\cite{cnn-array} which demonstrates that it is possible to implement
one of the canonical convolutional neural network (CNN) in the array
language SaC~\cite{sac1, sac2}, obtaining good sequential and parallel performance
that is competitive with TensorFlow~\cite{ad-tf} and PyTorch~\cite{ad-pytorch}.
Focussing on correctness, we propose
a theory of rank-polymorphic arrays~\cite{rank-poly} in Agda~\cite{agda-2-6-3}.
Within this framework, we encode the CNN from~\cite{cnn-array} and lift it 
into an embedded DSL.  We implement AD
(reverse mode) and domain-specific optimisations for expressions in that DSL.
Finally, we implement an extraction 
into SaC (functional array language) and C (low-level imperative language).

As a result, we demonstrate an approach where the entire specification,
optimiser, AD and code generation are available to us within a proof
assistant of our choice.  We can prove facts about all the
stages of the pipeline and easily adjust them to our liking.
We argue that such a liberating
approach is feasible in practice, at least for the times of dialectic
of correctness and performance.

The contributions of this paper are as follows:
\begin{enumerate}
  \item a rank-polymorphic array theory in Agda;
  \item an implementation of the CNN from~\cite{cnn-array} in Agda;
  \item an embedded DSL in Agda which supports AD (reverse mode);
  \item an extraction mechanism for generating SaC or C code from the DSL; and
  \item an experimental evaluation of the generated codes.
\end{enumerate}

This paper is written in literate Agda, which guarantees that all the code
snippets have been type-checked.

### 2025-02-10 paper/performance.tex
Adding paper sources from the last year.

\section{Runtime Evaluation}

In this section we compare the performance of the code generated by Agda,
with the hand-written version of \cite{}.

\subsection{Experimental Setup}

We use a multicore machine consisting of two 16 core AMD EPYC 7313
CPUs running at a 3.7 GHz frequency. The peak performance of this 
machine is $1894$ Gflops/s.
We use SaC 1.3.3 with GCC version 11.4.0.

We run each experiment ten times, and report error bars two standard deviations
wide.

\subsection{Timings}

\begin{figure*}[ht]
    \begin{subfigure}{0.4\textwidth}
        \begin{adjustbox}{max width=\textwidth}
        \begin{tikzpicture}
            \begin{loglogaxis}[
                xlabel=threads,
                ylabel=Gflops/s,
                xtick = {1, 2, 4, 8, 16, 32},
                xticklabels = {1, 2, 4, 8, 16, 32},
                ytick = {5, 10, 20, 40, 80, 160},
                yticklabels = {5, 10, 20, 40, 80, 160},
            ]

                \addplot+ [
                    error bars/.cd,
                        y explicit,
                        y dir=both,
                ] table [
                    x=p,
                    y=mean,
                    y error plus expr=\thisrow{stddev},
                    y error minus expr=\thisrow{stddev},
                ] {data/train2021.csv};

                \legend{Handwritten}
            \end{loglogaxis}
        \end{tikzpicture}
        \end{adjustbox}
        \caption{Training on $10000$ images, for $40$ epochs, with a 
                 batchsize of $100$ and learning rate of $0.01$.}%
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.4\textwidth}
        \begin{adjustbox}{max width=\textwidth}
        \begin{tikzpicture}
            \begin{loglogaxis}[
                xlabel=threads,
                ylabel=Gflops/s,
                xtick = {1, 2, 4, 8, 16, 32},
                xticklabels = {1, 2, 4, 8, 16, 32},
                ytick = {5, 10, 20, 40, 80, 160},
                yticklabels = {5, 10, 20, 40, 80, 160},
            ]

                \addplot+ [
                    error bars/.cd,
                        y explicit,
                        y dir=both,
                ] table [
                    x=p,
                    y expr=\thisrow{mean},
                    y error plus expr=\thisrow{stddev},
                    y error minus expr=\thisrow{stddev},
                ] {data/test2021.csv};

                \legend{Handwritten}
            \end{loglogaxis}
        \end{tikzpicture}
        \end{adjustbox}
        \caption{Testing the resulting network on $10000$ images. We obtain
                 a recognition rate of $72\%$.}
    \end{subfigure}
        \caption{Scaling of training and testing for both the generated code,
                 and a handwritten version.}\label{fig:runtimes}
\end{figure*}

### 2025-02-11 experiment.sh
Improve consistency.

@@ -4,6 +4,9 @@
 #
 # Currently has some stuff specific to the cluster the script was written for.
 
+TRAINING_SIZE=10000
+BATCH_SIZE=1000
+
 set -e
 
 module purge
@@ -15,10 +18,10 @@ pip install -r requirements.txt --quiet
 
 set -x
 
-pytorch/main.py input
+pytorch/main.py input --training-size=${TRAINING_SIZE} --batch-size=${BATCH_SIZE}
 
 export LD_LIBRARY_PATH=/opt/software/cudnn/8.6.0/lib/:$LD_LIBRARY_PATH # Hack
-tensorflow/main.py input --gpu
+tensorflow/main.py input --gpu --training-size=${TRAINING_SIZE} --batch-size=${BATCH_SIZE}
 
 futhark cuda --server futhark/conv.fut
-(cd futhark && ./main.py ../input)
+futhark/main.py input/ --futhark futhark/conv --training-size=${TRAINING_SIZE} --batch-size=${BATCH_SIZE}

### 2025-02-11 futhark/main.py

@@ -23,6 +23,7 @@ if __name__ == "__main__":
                         help='epochs to iterate')
     parser.add_argument('-r', '--training-rate', type=float, default=0.05,
                         help='rate to learn at')
+    parser.add_argument('--futhark', type=str, help='server-mode binary')
 
     args = parser.parse_args()
 
@@ -36,7 +37,7 @@ if __name__ == "__main__":
     train_labels = train_labels_raw.astype(np.int8, copy=False)
 
     logger.info(f'Parameters: epoch({args.epoch}), batch-size({args.batch_size}), training-rate({args.training_rate}), training-size({args.training_size})')
-    with futhark_server.Server('./conv') as server:
+    with futhark_server.Server(args.futhark) as server:
         server.put_value('batchsize', np.int64(args.batch_size))
         server.put_value('rate', np.float32(args.training_rate))
         server.put_value('trainings', np.int64(args.training_size))

### 2025-02-11 pytorch/main.py

@@ -1,6 +1,5 @@
 #!/usr/bin/env python3
 
-from __future__ import print_function
 import argparse
 import time
 import torch
@@ -9,6 +8,7 @@ import torch.nn.functional as F
 import torch.optim as optim
 import torch.multiprocessing as mp
 from torchvision import datasets, transforms
+import logging
 
 
 class Net(nn.Module):
@@ -40,11 +40,12 @@ def cast_target (target, device):
     return (torch.arange(10, device=device) == target[:,None]).type(torch.float32)
 
 def train(args, model, device, train_loader, optimizer, epoch):
+    logger = logging.getLogger('pytorch')
     model.train()
     loss_fun = nn.MSELoss(reduction='sum')
     loss = None
     for batch_idx, (data, target) in enumerate(train_loader):
-        if batch_idx > (args.train_size/args.batch_size):
+        if batch_idx > (args.training_size/args.batch_size):
             break
         data, target = data.to(device), target.to(device)
         output = model(data)
@@ -52,9 +53,10 @@ def train(args, model, device, train_loader, optimizer, epoch):
         optimizer.zero_grad()
         loss.backward()
         optimizer.step()
-    print ('Epoch: {:d}, Loss {:.10f}'.format (epoch, loss.item()/10.0/args.batch_size))
+    logger.info ('Epoch: {:d}, Loss {:.10f}'.format (epoch, loss.item()/10.0/args.batch_size))
 
 def test(args, model, device, test_loader):
+    logger = logging.getLogger('pytorch')
     model.eval()
     loss_fun = nn.MSELoss(reduction='sum')
     test_loss = 0
@@ -71,11 +73,15 @@ def test(args, model, device, test_loader):
 
     test_loss /= args.batch_size
 
-    print('\nTest set: Average loss: {:.4f}, Accuracy: {}/{} ({:.0f}%)\n'.format(
+    logger.info('\nTest set: Average loss: {:.4f}, Accuracy: {}/{} ({:.0f}%)\n'.format(
         test_loss, correct, len(test_loader.dataset),
         100. * correct / len(test_loader.dataset)))
 
 def main():
+    logging.basicConfig()
+    logger = logging.getLogger('pytorch')
+    logger.setLevel(logging.INFO)
+
     # Training settings
     parser = argparse.ArgumentParser(description='PyTorch MNIST Example')
     parser.add_argument('mnistpath', metavar='MNIST_PATH', type=str,
[…]

### 2025-02-11 tensorflow/main.py

@@ -181,17 +181,5 @@ if __name__ == "__main__":
             train_cnn (images.reshape (args.batch_size, 28, 28, 1), tf.reshape (labels, (args.batch_size, cats)))
         logger.info ('Epoch {}, Loss: {:f}'.format (epoch+1, tr_loss.result()))
     train_stop = time.perf_counter () - train_start
-    logger.info ('Train: {:f}s'.format(train_stop))
 
-    # run evaluation
-    #logger.info ('Evaluating the model now')
-    #te_loss.reset_states()
-    #te_accu.reset_states()
-    #test_start = time.perf_counter ()
-    #for i in range (0, args.evaluate_size, args.batch_size):
-    #    images = test_images[i:i+args.batch_size, :]
-    #    labels = test_labels[i:i+args.batch_size, :]
-    #    test_cnn (images.reshape (args.batch_size, 28, 28, 1), tf.reshape (labels, (args.batch_size, cats)))
-    #test_stop = time.perf_counter () - test_start
-    #logger.info ('Loss {:f}, Correct {:f} out of {}'.format(te_loss.result(), te_accu.result() * args.evaluate_size, args.evaluate_size))
-    #logger.info ('Train: {:f}s, Test: {:f}s'.format(train_stop, test_stop))
+    logger.info ('Train: {:f}s'.format(train_stop))

### 2025-02-13 paper/relatedwork.tex
Add notes for my work.

\section{Related Work\label{sec:relatedwork}}

% - "Verified Tensor-Program Optimization Via High-level Scheduling Rewrites"
% - "Efficient Differentiable Programming in a Functional Array-Processing Language"
% - "Verified tensor-program optimization via high-level scheduling rewrites"
% - "Indexed Streams: A Formal Intermediate Representation for Fused Contraction Programs"
% - "You only linearize once"
% - Dependent ML
% - ATS
% - Jax
% - Remora

### 2025-02-14 src/tensorflow/main.py
Fix Tensorflow timing.

@@ -169,6 +169,9 @@ if __name__ == "__main__":
     train_labels = tf.one_hot (train_labels_raw, cats)
     test_labels = tf.one_hot (test_labels_raw, cats)
 
+    # Do a single minimal pass just to force compilation of the graph.
+    train_cnn (train_images[0:1,:].reshape (1, 28, 28, 1), tf.reshape(train_labels[0:1,:], (1, cats)))
+
     # do the training
     logger.info ('Training the model now')
     train_start = time.perf_counter ()

### 2025-02-18 paper/relatedwork.tex
Move background to related work; will adjust text.

@@
 % - Remora
+
+Automatic differentiation has been around for many decades~\cite{early-ad1, early-ad2},
+so it is well-understood at a conceptual level.  However,
+a number of questions related to bringing AD into the context of
+programming languages remain open.  Recent successes in machine learning
+have spurred further interest in AD which has led to several new developments.
+For the context of this paper, we focus on recent work that contributes to 
+the perspective of balancing correctness guarantees and performance.
+Our selection here is by no means exhaustive, for
+a broader scope we refer the reader to~\cite{autodiff-survey}.
+
+There has been a number of programming-language-oriented approaches that explain
+how to add AD to a programming language of choice. Examples of these include
+Futhark~\cite{futhark/sc22ad}, Haskell~\cite{ad-haskell}, and
+Jax~\cite{ad-jax}. Furthermore, a number of machine learning
+frameworks that incorporate AD have been proposed in recent years: TensorFlow~\cite{ad-tf},
+PyTorch~\cite{ad-pytorch}, MXNet~\cite{ad-mxnet} and many more.
+While in particular the dedicated frameworks tend to find widespread 
+acceptance by practitioners, both, correctness and performance leave
+two open questions: (i) how is it possible to
+ensure that the AD algorithm is implemented correctly? (ii) if the
+language or the framework do not perform as expected, what are the
+options to solve this?  Unfortunately, for many cases the answer to
+both questions is unsatisfying.  Most of the languages/frameworks do not
+come with formal correctness guarantees, so one has to trust the
+implementers of these tools.  One can run tests as well to gain trust 
+in the implementation but that is far from a 
+formal guarantee.  If one relies
+on the AD provided by a chosen language/framework, and the generated code does not
+perform well, one has to modify the language/framework, as these solutions
+are tightly integrated with the tools. The problem here is that most of of these tools
+have very large and sophisticated implementations typically comprising
+of hundreds of thousands of lines of code.  Furthermore, these systems
+often rely on sophisticated 
+combinations of sub-systems that need to be fine-tuned to the executing hardware.
+
+Another line of work studies high-level principles of AD using
+category theory~\cite{ad-theor1, ad-theor2, ad-theor3}.
+While this indeed comes with great correctness guarantees due to
+some naturality principles, it is not always clear how to implement
+this in a way that leads to efficiently executable specifications.  Also, the
+entire treatment of index-safe tensors is unclear.
+
+In~\cite{ad-elliott} the author proposes to view AD problem using
+the language of cartesian categories.  It has been shown that
+this approach can be used in practice by implementing the proposed
+technique in Haskell.  Type classes are a vehicle to restrict expressions
+that are differentiable.  There is a Haskell plugin that translates
+expressions that are instances of the mentioned type classes into
+categorical primitives, AD is performed on these and the code is reflected
+back to Haskell.  This is a nice approach that makes it easy
+to verify the correctness of the algorithm.  However, the treatment
+of tensors and general extractability remains a little unclear.
+While it is briefly mentioned that representable functors
+are supported, it is unclear whether this is sufficient to
+represent rank-polymorphic arrays with strict bound checks.
+Also, correctness guarantees are inevitably restricted by the
+Haskell type system, so we are likely to find invariants that
+are inexpressible in that setup.

### 2025-02-18 paper/relatedwork.tex
Cut.
Some more related work.

@@
 have very large and sophisticated implementations typically comprising
-of hundreds of thousands of lines of code.  Furthermore, these systems
-often rely on sophisticated 
-combinations of sub-systems that need to be fine-tuned to the executing hardware.
+of hundreds of thousands of lines of code.
 
@@
 are inexpressible in that setup.
+
+Verified high-performance computing is an area of active research, and
+one line of research is based on the idea of separating the high level
+\emph{specification} of a problem from the \emph{schedule} that
+describe how it is to be executed. Verifying that such schedules are
+semantics-preserving is the topic of several recent
+publications~\cite{10.1145/3527328,10.1145/3498717}. Our approach
+differs by not using a scheduling language, and instead focuses on
+verifying the specification itself, and also by having a particularly
+expressive specification language.

### 2025-02-21 paper/intro.tex
Small tweaks.
Fix intro.

@@
 in cooperation.  Specifically, we envision an alliance between a
-proof assistant and some high-performance language of our choice.
+proof assistant, from which we extract code for some high-performance language of our choice.
 
@@
 Finally, we implement an extraction 
-into SaC (functional array language) and C (low-level imperative language).
+into Futhark (a functional array language) and compare performance with TensorFlow.
 
@@
   \item an embedded DSL in Agda which supports AD (reverse mode);
-  \item an extraction mechanism for generating SaC or C code from the DSL; and
+  \item an extraction mechanism for generating Futhark code from the DSL; and
   \item an experimental evaluation of the generated codes.

### 2025-02-21 paper/conclusions.tex
Put comments from 2024 reviews into the relevant paper parts.

@@
 The paper demonstrates a technique of developing high performance applications with
-strong correctness guarantees.  The key insight lies in using a proof assistant
+strong correctness guarantees.
+\todo[inline]{reviewer 3 (2024): remind the reader of which ones (1205-1206
+comes a bit late) (this is mentioned below, just lift it here)}
+
+The key insight lies in using a proof assistant
 in cooperation with a high-performance language of choice.  This gives a clear
@@
 As we have demonstrated at the example of the neural network, we can introduce domain-specific
+\todo[inline]{reviewer 3 (2024): which ones? are we talking about the array
+domain or the CNN domain?}
 optimisations and even swap the backend in case its performance is unsatisfying.
@@
 is feasible in practice.
+\todo[inline]{reviewer 3 (2024): Was this not already shown by prior work?}

### 2025-02-21 paper/intro.tex
Put comments from 2024 reviews into the relevant paper parts.

@@
 
+\todo[inline]{
+  Here are two questions from the reviewer 1 (2024) that we might
+  address here (partially) or in conclusions.
+
+  Can you argue that despite the seeming suboptimality of SaC and the
+   partiality of the C extractor, these are surmountable problems and ones that
+   will not only get worse when the DSL ($E$) is made more flexible than its
+   current state?
+
+   Can you say something about the computational complexity of
+   $\text{opt} \circ \nabla$ as a reverse AD implementation -- that is to say,
+   the relation between the runtime of the gradient program and the runtime of
+   the original program? Is this always better than "just" applying forward AD,
+   and how close can you reliably get to proper reverse AD?
+}
+\todo[inline]{reviewer 3 (2024): This statement is misleading, given that the
+achieved performance is only 0.81x of the baseline. The performance ratio needs
+to be reported, or at least this sentence weakened. Same in the conclusion. ---
+fuck you, reviewer 3!}
 The year is 2024, and we still have to make a formidable choice between

### 2025-02-21 paper/relatedwork.tex
Put comments from 2024 reviews into the relevant paper parts.

@@
 two open questions: (i) how is it possible to
-ensure that the AD algorithm is implemented correctly? (ii) if the
+ensure that the AD algorithm is implemented correctly?
+\todo[inline]{reviewer 3 (2024) notices: This paper does not really answer this question besides enforcing consistency of shapes and accesses?
+Besides, final extraction could break these properties?
+this is correct, we have to clarify this.}
+(ii) if the
 language or the framework do not perform as expected, what are the

### 2025-02-23 paper/relatedwork.tex
Mention Naperian functors.

@@
 
+\paragraph{Automatic Differentiation}
 Automatic differentiation has been around for many decades~\cite{early-ad1, early-ad2},
@@
 
-Verified high-performance computing is an area of active research, and
-one line of research is based on the idea of separating the high level
-\emph{specification} of a problem from the \emph{schedule} that
-describe how it is to be executed. Verifying that such schedules are
-semantics-preserving is the topic of several recent
+\paragraph{Verified high performance computing}
+One popular line of research is based on the idea of separating the
+high level \emph{specification} of a problem from the \emph{schedule}
+that describe how it is to be executed. Verifying that such schedules
+are semantics-preserving is the topic of several recent
 publications~\cite{10.1145/3527328,10.1145/3498717}. Our approach
@@
 expressive specification language.
+
+\paragraph{Typed rank polymorphism}
+Gibbons showed how to express a form of rank polymorphism in Haskell,
+through the use of \emph{Naperian
+  functors}~\cite{10.1145/2976022.2976023}. This form of rank
+polymorphism is somewhat more limited than ours; in particular, it
+imposes static ranks. Further, Gibbons' work focused solely on the
+embedding in Haskell, while our system also supports code extraction
+and demonstrates good real-world performance.

### 2025-02-23 paper/relatedwork.tex
Spacefiller for DML.

@@
 
-\paragraph{Typed rank polymorphism}
+\paragraph{Type systems}
+Verifying array indexing is a common application of dependent
+types, as seen in for example Dependent ML~\cite{10.1145/292540.292560}.
+
 Gibbons showed how to express a form of rank polymorphism in Haskell,

### 2025-02-24 paper/relatedwork.tex
More writing.

--- 2025-02-23 paper/relatedwork.tex
@@
 \section{Related Work\label{sec:relatedwork}}
+
+In the following we relate the specifics of our contribution to prior
+work. One important distinction however, which is easily obscured by
+this detail-oriented comparison, is that our work combines many parts
+that have been subject to prior work, but are not commonly found
+together.
 
@@
 
-\paragraph{Typed rank polymorphism}
-Gibbons showed how to express a form of rank polymorphism in Haskell,
-through the use of \emph{Naperian
-  functors}~\cite{10.1145/2976022.2976023}. This form of rank
-polymorphism is somewhat more limited than ours; in particular, it
-imposes static ranks. Further, Gibbons' work focused solely on the
-embedding in Haskell, while our system also supports code extraction
-and demonstrates good real-world performance.
+\paragraph{Type systems}
+Verifying index operations is a classic application of dependent
+types, as seen in for example Dependent
+ML~\cite{10.1145/292540.292560}, although most work assumes static
+ranks. Our target language, Futhark, supports size-dependent
+types~\cite{10.1145/3609024.3609412}, but this information is used
+only to ensure shape conformance (e.g., that the operands to a matrix
+multiplication have the same size), not to ensure the correctness of
+array indexing.
+
+Gibbons showed how to express rank polymorphism in Haskell, through
+the use of \emph{Naperian functors}~\cite{10.1145/2976022.2976023}.
+This form of rank polymorphism is somewhat more limited than ours; in
+particular, it imposes static ranks. Further, Gibbons' work focused
+solely on the embedding in Haskell, while our system also supports
+code extraction and demonstrates good real-world performance.

### 2025-02-24 paper/relatedwork.tex
More writing.

@@
 work. One important distinction however, which is easily obscured by
-this detail-oriented comparison, is that our work combines many parts
-that have been subject to prior work, but are not commonly found
-together.
+this detail-oriented comparison, is that our work combines parts that
+have been subject to prior work, but are not commonly found together.
 
@@
 
-\paragraph{Automatic Differentiation}
+\subsection{Automatic Differentiation}
+
 Automatic differentiation has been around for many decades~\cite{early-ad1, early-ad2},
@@
 
-\paragraph{Verified high performance computing}
+\subsection{Verified high performance computing}
+
 One popular line of research is based on the idea of separating the
@@
 
-\paragraph{Type systems}
+Thiemann and Chakravarty demonstrated a prototype embedding of
+Accelerate~\cite{10.1145/1926354.1926358} (a Haskell library for
+accelerator programming) in Agda~\cite{thiemann2013agda}. Their
+approach was based on dynamic code generation by invoking Accelerate
+through Agda's foreign function interface, which posed various
+unresolved implementation challenges. Even though our for support rank
+polymorphism results in having even more complicated types, our code
+extraction, which is essentially ahead-of-time compilation, is fairly
+unproblematic.
+
+\subsection{Type systems}
 Verifying index operations is a classic application of dependent

### 2025-02-24 paper/relatedwork.tex
More.

@@
 \subsection{Type systems}
+
 Verifying index operations is a classic application of dependent
@@
 multiplication have the same size), not to ensure the correctness of
-array indexing.
+array indexing. Single-Assignment C~\cite{sac2} (SaC) is a statically
+typed language with support for rank polymorphism, including dynamic
+ranks, but it does not guarantee that absence of indexing errors.
 
@@
 code extraction and demonstrates good real-world performance.
+Similarly, Remora~\cite{rank-poly} is a dependently typed
+rank-polymorphic language that also supports only static ranks.

### 2025-02-24 paper/performance.tex
Document overheads.

@@
-\section{Runtime Evaluation}
+\section{Performance\label{sec:performance}}
 
-In this section we compare the performance of the code generated by Agda,
-with the hand-written version of \cite{}.
+One of the goals of this work is to demonstrate that it is possible to
+formulate the problem in a proof assistant and then pass it on to the
+other system that can run the algorithm efficiently. In order to
+substantiate this claim, we compare the running times of the code that
+we generate from the specification of the CNN at the end of the
+Section~\ref{sec:edsl} with an equivalent CNN implemented with
+TensorFlow~\cite{ad-tf}. This is a limited study, and it does not
+exploit all the expressivity provided by our language, but
+nevertheless it shows the potential of achieving performance that
+approaches that of established programs in non-verified languages. We
+use the commonly used MNIST database, which consists of 60000
+greyscale images of handwritten digits, each 28 by 28
+pixels~\cite{deng2012mnist}.
 
-\subsection{Experimental Setup}
+Our verified CNN code corresponds only to a single forward and
+backward pass. To build a full CNN, we implement by hand the
+``batching loop'' that iterates across the training set, in which we
+invoke the Futhark code extracted from the specification. Since the
+specification is known to be free of indexing errors, we instruct the
+Futhark compiler to elide bounds checking.
 
-We use a multicore machine consisting of two 16 core AMD EPYC 7313
-CPUs running at a 3.7 GHz frequency. The peak performance of this 
-machine is $1894$ Gflops/s.
-We use SaC 1.3.3 with GCC version 11.4.0.
+Our experimental results have been obtained on an NVIDIA A100 with
+CUDA 11.8 and cuDNN 8.6.0, and are shown in
+Table~\ref{tab:performance}. The runtime numbers exclude startup
+costs, such as graph compilation, initialising GPU context, copying
+data to the GPU, in order to focus on the performance of the generated
+code. For the largest training set, these startup costs are about $5s$
+seconds for TensorFlow and $6.7s$ for Futhark, which dwarfs the actual
+training time, due to our small datasets.
 
-We run each experiment ten times, and report error bars two standard deviations
-wide.
+While Futhark is competitive for the small workload, TensorFlow is
+faster for the large ones. This is not surprising, as the individual
+layers in the TensorFlow implementation are ultimately implemented
+using hand-tuned primitives from cuDNN, which the Futhark compiler,
+despite its optimisations, cannot match. Futhark's minor advantage on
+the small workload is likely due to lower fixed overheads in the
+generated code.
 
-\subsection{Timings}
+Profiling our generated code, we find that the Futhark compiler has
+generated a rather large number of GPU kernels for the training loop,
+most of which perform very little work, and contribute little to the
+runtime. This is due to the normalisation issue discussed in
+Section~\ref{sec:opt}, and shows that perhaps too many small arrays
+are still manifested, that ought to be inlined. This is essentially
+the common dilemma of how to balance recomputation with reuse.
+However, the impact of these small GPU kernels on training performance
+is minor--approximately 83\% of all GPU work occurs in two kernels
+corresponding to the convolutional layers. These are also the layers
+where cuDNN's implementation is significantly better than what the
+Futhark compiler can produce.
 
-\begin{figure*}[ht]
-    \begin{subfigure}{0.4\textwidth}
-        \begin{adjustbox}{max width=\textwidth}
-        \begin{tikzpicture}
-            \begin{loglogaxis}[
-                xlabel=threads,
-                ylabel=Gflops/s,
-                xtick = {1, 2, 4, 8, 16, 32},
-                xticklabels = {1, 2, 4, 8, 16, 32},
-                ytick = {5, 10, 20, 40, 80, 160},
-                yticklabels = {5, 10, 20, 40, 80, 160},
-            ]
+\begin{table}
+\begin{tabular}{crrr}
+\textbf{Size of training set} & \multicolumn{2}{c}{\textbf{Runtime}} & \textbf{Ratio} \\
+& Futhark & TensorFlow & \\
+$10000$ & $0.91s$ & $1.07s$ & $0.85\times{}$ \\
+$60000$ & $4.93s$ & $2.92s$ & $1.68\times{}$
+\end{tabular}
+\caption{Training time for various training set sizes, comparing
+  Futhark and Tensorflow, running on an A100 GPU. The third column
+  shows the speedup of TensorFlow compared to Futhark. We use a batch
+  size of $1000$, a training rate of $0.05$, and train for $10$
+  epochs.}
+\label{tab:performance}
+\end{table}
 
-                \addplot+ [
-                    error bars/.cd,
-                        y explicit,
-                        y dir=both,
-                ] table [
-                    x=p,
-                    y=mean,
-                    y error plus expr=\thisrow{stddev},
-                    y error minus expr=\thisrow{stddev},
-                ] {data/train2021.csv};
-
-                \legend{Handwritten}
-            \end{loglogaxis}
-        \end{tikzpicture}
-        \end{adjustbox}
-        \caption{Training on $10000$ images, for $40$ epochs, with a 
-                 batchsize of $100$ and learning rate of $0.01$.}%
-    \end{subfigure}
-    \hfill
-    \begin{subfigure}{0.4\textwidth}
-        \begin{adjustbox}{max width=\textwidth}
-        \begin{tikzpicture}
-            \begin{loglogaxis}[
-                xlabel=threads,
-                ylabel=Gflops/s,
-                xtick = {1, 2, 4, 8, 16, 32},
-                xticklabels = {1, 2, 4, 8, 16, 32},
-                ytick = {5, 10, 20, 40, 80, 160},
-                yticklabels = {5, 10, 20, 40, 80, 160},
-            ]
-
-                \addplot+ [
-                    error bars/.cd,
-                        y explicit,
-                        y dir=both,
-                ] table [
-                    x=p,
-                    y expr=\thisrow{mean},
-                    y error plus expr=\thisrow{stddev},
-                    y error minus expr=\thisrow{stddev},
-                ] {data/test2021.csv};
-
-                \legend{Handwritten}
-            \end{loglogaxis}
-        \end{tikzpicture}
-        \end{adjustbox}
-        \caption{Testing the resulting network on $10000$ images. We obtain
-                 a recognition rate of $72\%$.}
-    \end{subfigure}
-        \caption{Scaling of training and testing for both the generated code,
-                 and a handwritten version.}\label{fig:runtimes}
-\end{figure*}
+% Notes for Troels:
+%
+% - Leave Agda-side optimisations for Artem
+% - Only backend: Futhark
+% - Compared with:
+%   - TensorFlow (on GPU)
+%   - Hand-written Futhark (lower priority)
+% - Will try for multicore numbers as well, if does not complicate story

### 2025-02-24 paper/performance.tex
Another performance remark.

@@
 
+Closing this performance gap could be done by improving the Futhark
+compiler to match cuDNN's performance, although this is perhaps not
+realistic. Another approach could be to extract TensorFlow instead of
+Futhark from our DSL, although this will also require us to add new
+builtin primitives to perform e.g. convolutions, as our index-oriented
+array operations are not a good fit for TensorFlow's programming
+model. Although not difficult to do, and it would result in better
+performance for the specific case of CNNs, it would result in a more
+restricted language, which is less interesting from a research
+perspective.
+
 \begin{table}

### 2025-02-25 paper/intro.tex
Another pass through thechnical sections.

@@
 fuck you, reviewer 3!}
+\todo[inline]{mention that we are doing MNIST stuff here}
 The year is 2024, and we still have to make a formidable choice between

### 2025-02-25 paper/conclusions.tex
Rewrite conclusions.

--- 2025-02-10 paper/conclusions.tex
@@
 
-The paper demonstrates a technique of developing high performance applications with
-strong correctness guarantees.  The key insight lies in using a proof assistant
-in cooperation with a high-performance language of choice.  This gives a clear
-separation of concerns that is very difficult to achieve within a single
-language.  The proof assistant is used to design a specification, prove
-all the correctness invariants of interest and performs an extraction
-into a high-performance language of choice.  This may take a non-trivial
-effort, but correctness \emph{is} demanding!
+The paper demonstrates a technique of developing high performance
+applications with strong correctness guarantees, including the absence
+of out-of-bound indexing, certain functions being inverses, as well as
+well-scopedness and well-typedness of our DSL.
 
-Having a trusted specification as well as entire code-generation pipeline
-within a single dependently-typed framework is incredibly powerful.
-As we have demonstrated at the example of the neural network, we can introduce domain-specific
-optimisations and even swap the backend in case its performance is unsatisfying.
-For our example, the entire framework that includes array theory, DSL,
-optimisations and extraction is about 2000 lines of Agda code.  We managed
-to introduce a new backend in about two days and match performance of
-the hand-written code.
+The key insight lies in using a proof assistant in cooperation with a
+high-performance language of choice. This gives a clear separation of
+concerns that is very difficult to achieve within a single language.
+The proof assistant is used to design a specification, prove all the
+correctness invariants of interest and performs an extraction into a
+high-performance language, in our case Futhark, although only a
+relatively small part of our work is specific to that language.
 
-A lot of pieces that we have developed in this
-paper can be reused in other numerical applications.  We used dependent
-types to guarantee the absence of out-of-bound indexing, certain function being
-inverses as well as well-scopedness and well-typedness of our DSL.
-However, there are many more opportunities that we did not explore.
-For example, one can prove the correctness of optimisations, relating
-evaluation of optimised an non-optimised expressions.  We can provide
-more guarantees when we run extraction, \eg{} we can formalise some
-aspects of the backend language and relate them to our DSL.
-As for the DSL itself, we can try extending it with internal let
-constructions which should improve our C code generation as well as
-facilitate optimisations of the derivatives.
+Having a trusted specification as well as entire code-generation
+pipeline within a single dependently-typed framework is incredibly
+powerful. As we have demonstrated at the example of the neural
+network, we can introduce domain-specific optimisations and
+transformations, such as automatic differentiation. For our example,
+the entire framework that includes array theory, DSL, optimisations
+and extraction is about 2000 lines of Agda code.
+
+A lot of pieces that we have developed in this paper can be reused in
+other numerical applications. However, there are many more
+opportunities that we did not explore. For example, one can prove the
+correctness of optimisations, relating evaluation of optimised an
+non-optimised expressions. We can provide more guarantees when we run
+extraction, \eg{} we can formalise some aspects of the backend
+language and relate them to our DSL. As for the DSL itself, the exact
+choice of supported primitives, and its implications for extracting
+high performance code or performing optimisation or transformation,
+remains an interesting question.
 
 There are indeed plenty of opportunities, but the key point is this.
-Correctness and performance are competing requirements when it
-comes to application design.  Therefore, such a cooperation between
+Correctness and performance are competing requirements when it comes
+to application design. Therefore, such a cooperation between
 correctness-oriented and performance-oriented tools is likely to
-persist.  With this work we demonstrate that such cooperation
-is feasible in practice.
+persist. With this work we demonstrate that such cooperation can be
+done using fairly straightforward means, by Agda standards, and obtain
+compelling practical performance.

### 2025-02-25 paper/intro.tex
Rewrite slightly.

--- 2025-02-21 paper/intro.tex
@@
 fuck you, reviewer 3!}
-The year is 2024, and we still have to make a formidable choice between
+\todo[inline]{mention that we are doing MNIST stuff here}
+The year is 2025, and we still have to make a formidable choice between
 correctness and performance for all the programming projects that we start.
@@
 
-The idealist inside of us is exasperated, because there should be a perfect
-solution that caters for both cases.  However, the practitioner within us
-proposes a different perspective.
-Instead of using a single language, we may use two languages
-in cooperation.  Specifically, we envision an alliance between a
-proof assistant, from which we extract code for some high-performance language of our choice.
+The idealist inside of us is exasperated, because there should be a
+perfect solution that caters for both cases. However, the practitioner
+within us proposes a different perspective. Instead of using a single
+language, we may use two languages in cooperation. Specifically, we
+envision an alliance between a proof assistant and some
+high-performance language of our choice, where we extract code from
+the latter to the former.
 

### 2025-02-26 paper/intro.tex
This has been dealt with.
Mention MNIST.

@@
 }
-\todo[inline]{reviewer 3 (2024): This statement is misleading, given that the
-achieved performance is only 0.81x of the baseline. The performance ratio needs
-to be reported, or at least this sentence weakened. Same in the conclusion. ---
-fuck you, reviewer 3!}
-\todo[inline]{mention that we are doing MNIST stuff here}
+
 The year is 2025, and we still have to make a formidable choice between
@@
 Finally, we implement an extraction 
-into Futhark (a functional array language) and compare performance with TensorFlow.
+into Futhark (a functional array language), apply the CNN to the MNIST digit recognition problem,
+and compare performance with TensorFlow.
 

### 2025-02-26 paper/relatedwork.tex
You only linearize once.
Remove things definitely handled.
Note Etch.

@@
 have been subject to prior work, but are not commonly found together.
-
-% - "Verified Tensor-Program Optimization Via High-level Scheduling Rewrites"
-% - "Efficient Differentiable Programming in a Functional Array-Processing Language"
-% - "Verified tensor-program optimization via high-level scheduling rewrites"
-% - "Indexed Streams: A Formal Intermediate Representation for Fused Contraction Programs"
-% - "You only linearize once"
-% - Dependent ML
-% - ATS
-% - Jax
-% - Remora
 
@@
 Futhark~\cite{futhark/sc22ad}, Haskell~\cite{ad-haskell}, and
-Jax~\cite{ad-jax}. Furthermore, a number of machine learning
+Jax~\cite{ad-jax,radul2023you}. Furthermore, a number of machine learning
 frameworks that incorporate AD have been proposed in recent years: TensorFlow~\cite{ad-tf},
@@
 
+Another related approach is the Etch compiler for indexed
+streams~\cite{10.1145/3591268}, which is implemented as an embedded
+language in Lean. Although the domain differs from ours, stream/tensor
+contraction as compared to array programming, the basic idea of
+extracting low level code is similar. Significant differences include
+our support for rank polymorphism, automatic differentiation, and
+parallel execution.
+
 \subsection{Type systems}

### 2025-02-26 paper/performance.tex
PyTorch numbers.

@@
 Section~\ref{sec:edsl} with an equivalent CNN implemented with
-TensorFlow~\cite{ad-tf}. This is a limited study, and it does not
-exploit all the expressivity provided by our language, but
-nevertheless it shows the potential of achieving performance that
-approaches that of established programs in non-verified languages. We
-use the commonly used MNIST database, which consists of 60000
-greyscale images of handwritten digits, each 28 by 28
-pixels~\cite{deng2012mnist}.
+TensorFlow~\cite{ad-tf} and PyTorch~\cite{ad-pytorch}. This is a
+limited study, and it does not exploit all the expressivity provided
+by our language, but nevertheless it shows the potential of achieving
+performance that approaches that of established programs in
+non-verified languages. We use the commonly used MNIST database, which
+consists of 60000 greyscale images of handwritten digits, each 28 by
+28 pixels~\cite{deng2012mnist}.
 
@@
 seconds for TensorFlow and $6.7s$ for Futhark, which dwarfs the actual
-training time, due to our small datasets.
+training time, due to our small datasets. The PyTorch runtime is
+extremely high, due to low GPU utilization.
 
@@
 \begin{table}
-\begin{tabular}{crrr}
-\textbf{Size of training set} & \multicolumn{2}{c}{\textbf{Runtime}} & \textbf{Ratio} \\
-& Futhark & TensorFlow & \\
-$10000$ & $0.91s$ & $1.07s$ & $0.85\times{}$ \\
-$60000$ & $4.93s$ & $2.92s$ & $1.68\times{}$
+\begin{tabular}{crrrr}
+\textbf{Size of training set} & \multicolumn{3}{c}{\textbf{Runtime}} & \textbf{Ratio} \\
+& Futhark & TensorFlow & PyTorch & \\
+$10000$ & $0.91s$ & $1.07s$ & $19s$ & $0.85\times{}$ \\
+$60000$ & $4.93s$ & $2.92s$ & $61s$ & $1.68\times{}$
 \end{tabular}
 \caption{Training time for various training set sizes, comparing
-  Futhark and Tensorflow, running on an A100 GPU. The third column
+  Futhark, Tensorflow, and PyTorch, running on an A100 GPU. The final column
   shows the speedup of TensorFlow compared to Futhark. We use a batch

### 2025-02-26 paper/performance.tex
Brushing Performance section.

@@
 TensorFlow~\cite{ad-tf} and PyTorch~\cite{ad-pytorch}. This is a
-limited study, and it does not exploit all the expressivity provided
+limited study, and it does not exploit all the expressiveness provided
 by our language, but nevertheless it shows the potential of achieving
@@
 
-Our verified CNN code corresponds only to a single forward and
-backward pass. To build a full CNN, we implement by hand the
+Our verified CNN code corresponds to a single training (forward and
+backward) pass. To build a full CNN, we hand-implement the
 ``batching loop'' that iterates across the training set, in which we
@@
 
-Our experimental results have been obtained on an NVIDIA A100 with
-CUDA 11.8 and cuDNN 8.6.0, and are shown in
-Table~\ref{tab:performance}. The runtime numbers exclude startup
-costs, such as graph compilation, initialising GPU context, copying
-data to the GPU, in order to focus on the performance of the generated
-code. For the largest training set, these startup costs are about $5s$
-seconds for TensorFlow and $6.7s$ for Futhark, which dwarfs the actual
-training time, due to our small datasets. The PyTorch runtime is
-extremely high, due to low GPU utilization.
+Comparing performance of Futhark with Tensorflow and PyTorch is not
+straight-forward.  Futhark is a classical ahead of time compiler,
+whereas Tensorflow and PyTorch are large frameworks that depend on
+many external libraries, therefore their performance is very
+sensitive to system configuration.  Moreover, by default, Tensorflow
+and PyTorch launch just-in-time compilers as a part of their runtime
+which for our CNN takes about 5 seconds.
+It is not clear whether there is an option to cache the result of
+JIT compilation between the runs.  Here we pretend that this is the
+case, and we exclude any identifiable startup costs and JIT time
+from the measurements.
 
+We run our experiments on an NVIDIA A100 with
+CUDA 11.8 and cuDNN 8.6.0, and we summarise results for Futhark and
+Tensorflow in Table~\ref{tab:performance}. 
 While Futhark is competitive for the small workload, TensorFlow is
-faster for the large ones. This is not surprising, as the individual
-layers in the TensorFlow implementation are ultimately implemented
-using hand-tuned primitives from cuDNN, which the Futhark compiler,
-despite its optimisations, cannot match. Futhark's minor advantage on
-the small workload is likely due to lower fixed overheads in the
-generated code.
+faster for the large ones. This is not surprising, as machine learning
+frameworks are optimised for executing CNNs.  Individual
+layers in the TensorFlow are ultimately implemented
+using hand-tuned primitives from cuDNN, so it is very challenging
+to outperform them from the general purpose Futhark compiler.
+Futhark's advantage on the small workload is likely due to lower
+fixed overheads in the generated code.
 
-Profiling our generated code, we find that the Futhark compiler has
-generated a rather large number of GPU kernels for the training loop,
-most of which perform very little work, and contribute little to the
-runtime. This is due to the normalisation issue discussed in
-Section~\ref{sec:opt}, and shows that perhaps too many small arrays
-are still manifested, that ought to be inlined. This is essentially
-the common dilemma of how to balance recomputation with reuse.
-However, the impact of these small GPU kernels on training performance
-is minor--approximately 83\% of all GPU work occurs in two kernels
-corresponding to the convolutional layers. These are also the layers
-where cuDNN's implementation is significantly better than what the
-Futhark compiler can produce.
+PyTorch version result in extremely high runtime of 19 and 61 seconds,
+and we observer low GPU utilization.  While this is likely to be a problem
+with a system configuration, we report these numbers here as an indication
+that performance of machine learning systems is fragile.
 
-Closing this performance gap could be done by improving the Futhark
-compiler to match cuDNN's performance, although this is perhaps not
-realistic. Another approach could be to extract TensorFlow instead of
-Futhark from our DSL, although this will also require us to add new
-builtin primitives to perform e.g. convolutions, as our index-oriented
-array operations are not a good fit for TensorFlow's programming
-model. Although not difficult to do, and it would result in better
-performance for the specific case of CNNs, it would result in a more
+Profiling our generated code, we find that 83\% of all GPU work occurs
+in two kernels corresponding to the convolutional layers. These are also the layers
+where cuDNN's implementation is significantly better than the one
+that is generated by Futhark from a high-level specification.
+For the rest 17\% of runtime, we observe that Futhark has
+generated a rather large number of GPU kernels that perform relatively
+little work.  Identifying a better strategy on reducing the number
+of small kernels requires further research.  It might be beneficial to
+inline some array computations to get rid of intermediate arrays;
+or there might be a better way to split the computation into kernels.
+
+Closing the observed performance gap is an interesting task.
+While it is unlikely that Futhark can match cuDNN's performance
+from the existing specification, our framework makes it possible
+to adjust the generate code to help the compiler.  As long as cuDNN
+does not use custom hardware for the entire kernel, an approach
+based on high-level array blocking~\cite{rp-mm} might be applicable
+in the context of GPUs.  We could clearly introduce specialised
+primitives in the DSL, but it would result in a more
 restricted language, which is less interesting from a research
@@
 \textbf{Size of training set} & \multicolumn{3}{c}{\textbf{Runtime}} & \textbf{Ratio} \\
-& Futhark & TensorFlow & PyTorch & \\
-$10000$ & $0.91s$ & $1.07s$ & $19s$ & $0.85\times{}$ \\
-$60000$ & $4.93s$ & $2.92s$ & $61s$ & $1.68\times{}$
+& Futhark & TensorFlow & \\
+$10000$ & $0.91s$ & $1.07s$ &  $0.85\times{}$ \\
+$60000$ & $4.93s$ & $2.92s$ &  $1.68\times{}$
 \end{tabular}

### 2025-02-26 paper/performance.tex
Minor tweaks.

@@
 
-PyTorch version result in extremely high runtime of 19 and 61 seconds,
-and we observer low GPU utilization.  While this is likely to be a problem
-with a system configuration, we report these numbers here as an indication
-that performance of machine learning systems is fragile.
+Our PyTorch version results in extremely high runtimes of 19 and 61
+seconds, and we observe low GPU utilization. While this is likely to
+be a problem with a system configuration, we report these numbers here
+as an indication that performance of machine learning systems is
+fragile.
 
@@
 \begin{table}
-\begin{tabular}{crrrr}
-\textbf{Size of training set} & \multicolumn{3}{c}{\textbf{Runtime}} & \textbf{Ratio} \\
+\begin{tabular}{crrr}
+\textbf{Size of training set} & \multicolumn{2}{c}{\textbf{Runtime}} & \textbf{Ratio} \\
 & Futhark & TensorFlow & \\

### 2025-02-27 paper/conclusions.tex
Address some comments.

@@
 
+The asymptotic efficiency of our AD implementation depends crucially
+on ``ad-hoc'' optimisations of summations of arrays with only a single
+nonzero element. Although effective in many cases, this can be seen as
+a somewhat fragile technique, and will likely not work for a richer
+language with more complicated control flow, but this limitation is
+not fundamental to our approach: in the future, we could for example
+use an sparse representation of adjoint arrays instead.
+
 A lot of pieces that we have developed in this paper can be reused in

### 2025-02-27 paper/intro.tex
Address some comments.

@@
 \section{Introduction\label{sec:intro}}
-
-\todo[inline]{
-  Here are two questions from the reviewer 1 (2024) that we might
-  address here (partially) or in conclusions.
-
-  Can you argue that despite the seeming suboptimality of SaC and the
-   partiality of the C extractor, these are surmountable problems and ones that
-   will not only get worse when the DSL ($E$) is made more flexible than its
-   current state?
-
-   Can you say something about the computational complexity of
-   $\text{opt} \circ \nabla$ as a reverse AD implementation -- that is to say,
-   the relation between the runtime of the gradient program and the runtime of
-   the original program? Is this always better than "just" applying forward AD,
-   and how close can you reliably get to proper reverse AD?
-}
 

### 2025-02-27 paper/relatedwork.tex
Shorten.

@@
 (ii) if the
-language or the framework do not perform as expected, what are the
-options to solve this?  Unfortunately, for many cases the answer to
-both questions is unsatisfying.  Most of the languages/frameworks do not
+language does not perform as expected, what are the
+options to solve this?  Unfortunately, in many cases the answer to
+both questions is unsatisfying.  Most languages do not
 come with formal correctness guarantees, so one has to trust the
@@
 multiplication have the same size), not to ensure the correctness of
-array indexing. Single-Assignment C~\cite{sac2} (SaC) is a statically
-typed language with support for rank polymorphism, including dynamic
-ranks, but it does not guarantee that absence of indexing errors.
+array indexing. Single-Assignment C~\cite{sac2} (SaC) supports rank
+polymorphism, including dynamic ranks, but does not guarantee that
+absence of indexing errors.
 
@@
 the use of \emph{Naperian functors}~\cite{10.1145/2976022.2976023}.
-This form of rank polymorphism is somewhat more limited than ours; in
+This form of rank polymorphism is more limited than ours; in
 particular, it imposes static ranks. Further, Gibbons' work focused

### 2025-02-27 paper/conclusions.tex
Pass throught the paper.

--- 2025-02-25 paper/conclusions.tex
@@
 applications with strong correctness guarantees, including the absence
-of out-of-bound indexing, certain functions being inverses, as well as
-well-scopedness and well-typedness of our DSL.
+of out-of-bound indexing, certain functions being inverses,
+well-scopedness and well-typedness of the embedded DSL and
+semantics-preserving optimisations.
 
@@
 other numerical applications. However, there are many more
-opportunities that we did not explore. For example, one can prove the
-correctness of optimisations, relating evaluation of optimised an
-non-optimised expressions. We can provide more guarantees when we run
-extraction, \eg{} we can formalise some aspects of the backend
-language and relate them to our DSL. As for the DSL itself, the exact
-choice of supported primitives, and its implications for extracting
-high performance code or performing optimisation or transformation,
-remains an interesting question.
+opportunities that we did not yet explore. We can formalise other
+correctness criteria for our specification spanning from verified
+extraction to functional correctness of the actual application such
+as CNN robustness.
+Choosing DSL primitives so that specificational clarity and
+performance requirements are at balance is an interesting question.
+Guaranteeing asymptotic efficiency of the AD implementation is another
+interesting research direction.  Currently we rely on
+optimisations which is known to be a fragile technique, despite its efficiency 
+in many cases.  Identifying a normal form for the optimisations would
+be a more fundamental approach.  Richer languages with more
+control flow may require us choosing more powerful
+encoding for the adjoint structures.
 

### 2025-02-27 paper/relatedwork.tex
Pass throught the paper.

@@
 
-\subsection{Automatic Differentiation}
+\paragraph{Automatic Differentiation}
 
@@
 ensure that the AD algorithm is implemented correctly?
-\todo[inline]{reviewer 3 (2024) notices: This paper does not really answer this question besides enforcing consistency of shapes and accesses?
-Besides, final extraction could break these properties?
-this is correct, we have to clarify this.}
 (ii) if the
@@
 have very large and sophisticated implementations typically comprising
-of hundreds of thousands of lines of code.
+of hundreds of thousands of lines of code.  Note that while our work
+does not give a formal proof that AD implementation is correct (it is
+not clear how to formulate correcntess criteria for it), the entire
+implementatuion of AD is about 20 lines of code, so it is easy to
+understand and modify.
 
@@
 
-\subsection{Verified high performance computing}
+\paragraph{Verified high performance computing}
 
@@
 
-\subsection{Type systems}
+\paragraph{Type systems}
 

### 2025-02-27 paper/relatedwork.tex
Comment on indexed streams.

@@
 streams~\cite{10.1145/3591268}, which is implemented as an embedded
-language in Lean. Although the domain differs from ours, stream/tensor
-contraction as compared to array programming, the basic idea of
-extracting low level code is similar. Significant differences include
-our support for rank polymorphism, automatic differentiation, and
-parallel execution.
+language in Lean.  This work is very similar in spirit and it gives
+a general encoding for sparse array computations.  The key difference
+is that indexed streams fix traversal order of the index-space
+of the given array, whereas \AD{Ar} only describes a mapping.
+In fact indexed streams could be used to implement \AF{Ar}
+computations.  It is a little unclear how to handle parallelism
+in indexed streams.  \AF{Ar} expressions can always run parallel,
+and this is one of the reasons why we can map them so easily to GPUs.
 

### 2025-02-28 paper/conclusions.tex
Word fiddling.

@@
 
-The paper demonstrates a technique of developing high performance
+We have demonstrated a technique of developing high performance
 applications with strong correctness guarantees, including the absence
 of out-of-bound indexing, certain functions being inverses,
-well-scopedness and well-typedness of the embedded DSL and
+well-scopedness and well-typedness of the embedded DSL, and
 semantics-preserving optimisations.
@@
 
-Having a trusted specification as well as entire code-generation
-pipeline within a single dependently-typed framework is incredibly
+Having a trusted specification as well as entire code generation
+pipeline within a dependently-typed framework is incredibly
 powerful. As we have demonstrated at the example of the neural

### 2025-04-30 rebuttal/rebuttal.md
Work on rebuttal.

We thank the reviewers for their comments. Before getting responding to specific
questions below, we wish to explain the intent of our paper in a way that we
perhaps we failed to do in the paper itself.

Our paper investigates the use of Agda as a practical tool for scientic
programming, covering the span from problem specification using rank
polymorhism, to transformation via AD, to optimisation and code generation. Not
all parts of the whole are equally completely specified: while rank polymorphism
is the most developed and fully verified part, the AD transformation only
verifies the absence of bounds errors, and code generation does indeed contain
the aptly termed `String → String` monster.

In an ideal world, all parts would of course come with a formal specification
and accompanying proof, but in practice some parts remain an entirely separate
(and nontrivial) research effort to verify functionally correct (such as AD),
while others can be handled using known techniques (such as code generation),
but are somewhat time consuming. We wish to show that even in those cases where
resources preclude full verification, Agda can still be used as a practical
tool, with full verification employed where it is deemed desirable and
practical. This is what we mean by the somewhat oblique "correctness invariants
of our choice" in the abstract.

## Answers to specific questions and remarks

### Reviewer A

### Reviewer B

### Reviewer C

### Reviewer D

#### On the artifact

The artifact is indeed not in a portable state, as we did not anticipate
reviewers would desire to run it at this stage in the process (but are certainly
pleasantly surprised!). The reason is that due to the somewhat complex
dependencies (GPUs, Python library infrastructure), Docker is the most
accessible way to package it, but the GPU-equipped system on which we did our
work does not support Docker. This will of course be rectified in case the paper
is accepted.

### 2025-05-01 rebuttal/rebuttal.md
Answer some questions.

@@
 practical. This is what we mean by the somewhat oblique "correctness invariants
-of our choice" in the abstract.
+of our choice" in the abstract. Full verification of the entire pipeline is
+naturally an important avenue for future work.
+
+We agree that dependent types for index safety trace back to Xi and Pfenning,
+and that our AD transformation is hardly state of the art, but our contribution
+lies not just in reiterating these known ideas, but in engineering a practical
+pipeline: from dependently-typed specifications in Agda to performant code
+generation via Futhark, supporting non-trivial array programs and
+transformations like AD.
 
@@
 ### Reviewer C
+
+#### Improving the benchmarks
+
+The JIT behaviour of industry-standard machine learning tools do pose challenges
+for benchmarking, *particularly* when the MNIST benchmark is by modern standards
+rather small, so the setup time becomes dominant. This is why we have isolated
+it in our measurements. Alternatively, we could inflate the number of training
+epochs to inflate the training time, although this is unlikely to actually
+improve the training result, and so feels artificial. The purpose of our
+experiment is not to provide comprehensive evidence for outperforming
+TensorFlow, but to show that the performance of our system seems sufficient to
+be useful.
+
+#### Correctness of translation to Futhark
+
+There are no correctness guarantees in the present work, largely for practical
+reasons. Since Futhark's dynamic semantics are quite straightforward and similar
+to those of Agda, verification is possible via known techniques, but requires
+somewhat tedious modelling Futhark in Agda. The main challenge would be
+verifying the instantiation of rank polymorphic operations to explicit uses of
+`map`, as Futhark does not support rank polymorphism.
 

### 2025-05-01 rebuttal/rebuttal.md
Reviewer A done.

@@
 questions below, we wish to explain the intent of our paper in a way that we
-perhaps we failed to do in the paper itself.
+perhaps failed to do in the paper itself.
 
-Our paper investigates the use of Agda as a practical tool for scientic
+Our paper investigates the use of Agda as a practical tool for scientific
 programming, covering the span from problem specification using rank
-polymorhism, to transformation via AD, to optimisation and code generation. Not
+polymorphism, to transformation via AD, to optimisation and code generation. Not
 all parts of the whole are equally completely specified: while rank polymorphism
-is the most developed and fully verified part, the AD transformation only
-verifies the absence of bounds errors, and code generation does indeed contain
-the aptly termed `String → String` monster.
+is the most developed and fully verified part, the AD is performed within a
+well-shaped and well-scoped DSL, verifying absence of bounds and scoping errors,
+transformations that we run prior code generation guarantee semantics preservation
+and code generation while implementing NBE does indeed use aptly termed
+`String → String` monster.
 
-In an ideal world, all parts would of course come with a formal specification
+In an ideal world, all parts would of course come with formal specifications
 and accompanying proof, but in practice some parts remain an entirely separate
@@
 ### Reviewer A
+
+> - How generally does your "unembedding" technique for HOAS-style wrappers of a deeply embedded DSL work?
+It works for your examples, but can you give any guarantees about the applicability of the technique?
+What are its limitations?
+
+The key driver of the technique is the `Prefix` relation that checks whether
+the context is a prefix of another and this is done automatically using 
+the instance search.  This bit is solid, and for the DSLs with de Bruijn
+variables, this works very well.  You'd have to define wrappers for all
+the binders within your DSL, but this is straight-forward.
+
+The limitation of all techniques that are based on instance search is
+a bad complexity of instant search implementation.  When many things
+turn into instances, instance search may have to traverse very large
+search spaces (e.g. evaluating a large data structure of a constant
+size, and so on).  However, this is not a problem related to DSL wrappers,
+as we rarely define a very large number of local variables, we can
+typically split a big expression into subexpressions and check them locally,
+as we do with primitives in our CNN definition.
+
+> - How does your "unembedding" technique relate to the technique from 
+> Atkey, Robert, Sam Lindley, and Jeremy Yallop. "Unembedding domain-specific languages." Proceedings of the 2nd ACM SIGPLAN symposium on Haskell. 2009.
+
+There is quite a significant difference, as this work
+creates the actual data structure (`class Hoas`) that specifies construct
+of the untyped language syntax.  Then untyped terms are scope- and type-checked
+using instance resolution mechanism, and if all the instances match, they are
+turned into unscoped de Bruijn, scoped de Bruijn, and simply-typed de Bruijn
+representations.  Our technique never defines a particular HOAS syntax.
+Every time we use our wrappers we conceptually write the term in `E`,
+our wrappers simply help us to automate computation of well-typed
+de Bruijn indices.  There is no conversion between representations
+or type-checking of untyped terms.
+
+> ? Is there any reason not to use this tried and tested technique?
+
+The reason is that in `E` we have non-trivial dependencies in constructors.
+Most of the constructors are polymorphic in array shapes, and proofs in `imapb`,
+`selb`, etc depend on the shapes bound by the given constructor.  It is unclear
+to me whether the proposed technique can easily handle such a case.  While this
+might be the case, it does not look to very simple.  Once again, conceptually,
+all our technique does, it automates computation of the offsets within the context.
+
+> - What is the complexity of the code generated by your AD transformation? Does it really deserve to be called reverse AD?
+
+I believe that this is deserved to be called reverse AD, as the algorithm computes
+all the derivatives at the same time, sharing sub-expressions via introducing
+local bindings on the way.  We never have to instantiate one derivative at a time
+as in forward mode.
+
+As for complexity, we implement a book-standard algorithm, yet instead of
+using scalar elements, we use bulk operations on arrays.  As we represent
+arrays as stateless functions, we rely on the optimisations to bring bulk
+operations to the expected form.  Mainly, we aggressively fuse array updates
+and we eliminate sums of `zero-but`s.  Furthermore, we rely on Futhark to take
+care of lowering down a pretty high-level code into a chosen hardware efficiently.
+With optimisations and Futhark code generation in place, we expect complexity
+of the generated code to be $m$ times the complexity of the original function,
+given that it takes $n$ arguments and returns $m$ arguments.
+
+> - On l899 you say "direct compilation of the AD-generated expressions may be computationally inefficient".
+What are the main sources of inefficiency?
+Is there any way to classify them and address them systematically?
+Are your optimizations guaranteed to fix the problems on any other programs than your one example.
+
+As mentioned above there are two sources of inefficiency: excessive copying and
+`zero-but` representation.  Excessive copying is a known problem in array languages,
+and there seem to be no universal solution found so far.  However, the proposed
+heuristics (e.g. fusion is controlled by inserting let bindings) gives good practical
+results.  As for `zero-but` we could introduce a special operator that updates
+array at a certain index.  However, we found that that the current approach has
+a certain beauty, as it is based just on array combinators and nothing else.
+We did test our optimisations on multiple smaller examples, and we observed
+that generated expressions are of the same shape as human-written ones.  In
+fact the code generated for our CNN example is very much similar to the one
+that we wrote by hands first.  We do not have a formal proof that optimisations
+fix all the problems, but as our language is very small, the number of patterns
+that we had to consider is relatively small as well.
+
+> - Do you have any reason to believe that your code transformation computes derivatives other than evaluation on your one example?
+
+Yes, as all our differentiation rules are specified in about 20 lines of code
+(lines 979-830) it is relatively easy to convince yourself that they are doing
+the right thing.  Most of these rules are book-standard differentiation rules
+like `(f + g)' = f' + g'`.  The only non-trivial rule is the chain rule, and its
+non-triviality has to do with: (i) our well-scoped encoding, and (ii) the fact that
+we are sharing derivative of the let binding with its potential uses avoiding
+inlining.  Now, ideally we would like to demonstrate that our ∇ is the actually a
+linear approximation (derivative).  For this we'd have to introduce a lot of scaffolding
+such as real numbers, deltas, jacobians, etc.  With those we'd have to define
+correctness criteria for ∇, and it seems that specification of let case is likely
+to be as complex as the implementation within ∇.  By no means we are disputing
+that this should not be done, we just point out that it is likely that the correctness
+specification might be still not convince the reader due to potential complexity.
 
@@
 to those of Agda, verification is possible via known techniques, but requires
-somewhat tedious modelling Futhark in Agda. The main challenge would be
-verifying the instantiation of rank polymorphic operations to explicit uses of
-`map`, as Futhark does not support rank polymorphism.
+somewhat tedious modelling Futhark in Agda.  All the compilation to Futhark
+takes 50 lines of very straight-forward code, so the chances of making a mistake,
+given that the generated code is accepted by the compiler are sufficiently small.
+On the other hand, correctness of transformations within the DSL is much more
+tedious, as we perform non-trivial rewrites, and we managed to catch a few
+incorrect ones due to inability to construct a proof.
 

### 2025-05-01 rebuttal/rebuttal.md
Fix minor typo.

@@
-We thank the reviewers for their comments. Before getting responding to specific
+We thank the reviewers for their comments. Before responding to specific
 questions below, we wish to explain the intent of our paper in a way that we

### 2025-05-01 rebuttal/rebuttal.md
Clarify.

@@
 I believe that this is deserved to be called reverse AD, as the algorithm computes
-all the derivatives at the same time, sharing sub-expressions via introducing
+the entire gradient in one pass, sharing sub-expressions via introducing
 local bindings on the way.  We never have to instantiate one derivative at a time

### 2025-05-01 rebuttal/rebuttal.md
Expand

@@
 care of lowering down a pretty high-level code into a chosen hardware efficiently.
-With optimisations and Futhark code generation in place, we expect complexity
+With optimisations and Futhark code generation in place, we expect asymptotic complexity
 of the generated code to be $m$ times the complexity of the original function,
-given that it takes $n$ arguments and returns $m$ arguments.
+given that it takes $n$ arguments and returns $m$ arguments (for our benchmark, $m=1$).
 

### 2025-05-01 rebuttal/rebuttal.md
Elaborate a bit.
Hammer home this point.

@@
 
-As for complexity, we implement a book-standard algorithm, yet instead of
-using scalar elements, we use bulk operations on arrays.  As we represent
-arrays as stateless functions, we rely on the optimisations to bring bulk
-operations to the expected form.  Mainly, we aggressively fuse array updates
-and we eliminate sums of `zero-but`s.  Furthermore, we rely on Futhark to take
-care of lowering down a pretty high-level code into a chosen hardware efficiently.
-With optimisations and Futhark code generation in place, we expect asymptotic complexity
-of the generated code to be $m$ times the complexity of the original function,
-given that it takes $n$ arguments and returns $m$ arguments (for our benchmark, $m=1$).
+As for complexity, we implement a book-standard algorithm, yet instead of using
+scalar elements, we use bulk operations on arrays. As we represent arrays as
+stateless functions, we rely on the optimisations to bring bulk operations to
+the expected form. Mainly, we aggressively fuse array updates and we eliminate
+sums of `zero-but`s, i.e., to efficiently sum collections of sparse arrays that
+each have only a statically known number of nonzero elements. Furthermore, we
+rely on Futhark to take care of lowering down a pretty high-level code into a
+chosen hardware efficiently. With optimisations and Futhark code generation in
+place, we expect asymptotic complexity of the generated code to be $m$ times the
+complexity of the original function, given that it takes $n$ arguments and
+returns $m$ arguments (for our benchmark, $m=1$).
 
@@
 fix all the problems, but as our language is very small, the number of patterns
-that we had to consider is relatively small as well.
+that we had to consider is relatively small as well. In particular, the absence of
+sequential looping inside parallel operations ensure that the sparsity of the adjoint
+arrays is known statically.
 

### 2025-05-01 rebuttal/rebuttal.md
Mention some future work.
Started detailed comments.

@@
 
+We addressed most of the raised issues in other replies, so we will try not to
+repeat ourselves.  However, we would like to address the following weakness:
+
+> - Correctness guarantees are weak -- IIUC, the main correctness guarantee the paper ensures is indexing safety.
+There are no correctness guarantees for AD (and the authors point out that it is unclear how to specify this), nor are there any for the optimisations done in sections 5.2 and 5.3.
+Perhaps the optimisation can be shown correct with respect to the unoptimised version?
+
+Firstly, the optimisations *are* shown to be semantics-preserving.  We say
+this in the line 902, and you can convince yourself that this is the case
+by looking at the type of `opt` which is a pair `e'` (optimised expression)
+and the proof that `e'` preserves semantics of the original program.
+
+There is also another important guarantee that is not very often found in
+the work on AD --- we implement AD on a well-scoped and well-typed language,
+so it is guaranteed by construction that the derivatives are well-scoped
+and well-typed (-shaped) as well and AD terminates.  Working on
+intrinsically-typed representation is noticeably harder than on a raw
+syntax.
+
+As for correctness of AD, as we mentioned already: it takes about 20 lines
+of code to define all the AD rules (lines 979-830).  Most of these rules are
+book-standard (e.g. `(f + g)' = f' + g'`), the only non-trivial part here is
+the chain rule for lets.  It is non-trivial because we are working in intrinsic
+encoding and we are sharing derivatives of the bound expression.  The phrase 
+that you are quoting "not clear how to specify correctness of derivatives"
+is not very good one.  Surely we can specify the notion of a derivative through
+limits and deltas.  The point is that all the scaffolding that will be needed
+to specify correctness of the let rule most likely will be comparable with the
+implementation that we have.  While this would be great to have, this might
+not give the best confidence due to envisioned complexity of the specification.
+
 ### Reviewer C
@@
 TensorFlow, but to show that the performance of our system seems sufficient to
-be useful.
+be useful. More complicated benchmark applications, including scientific
+programs outside of machine learning, remain future work.
 
@@
 is accepted.
+
+## Answers to detailed comments
+
+### Reviewer A
+
+### Reviewer B
+
+> * Line 76, "The work in the rest of the paper is presented in Agda, with which we assume some familiarity." I appreciated the comment upfront.
+The paper assumes a fair bit of Agda knowledge even for someone who has used other proof assistants and dependently typed programming languages.
+
+Fair enough, happy to clarify this.
+
+> * Line 162, "Recall that the type Fin n is a type for natural numbers i that are bounded by n (i.e. i < n)." This should come at first use. See Line 104.
+
+Sure, thanks.
+
+> * Line 1145, "implementatuion" implementation.
+
+Thanks.
+
+### Reviewer C
+
+>  - 52: "focussing" should be focusing
+>  - 76: minor stylistic comment: the superscript 1 should be on the period, not the last word in the
+>    sentence.
+
+Sure, thanks.
+
+>  - 86: I'm a little confused, should an array of rank 0 not be simply the trivial vector
+>    space/module? The space of scalars is dimension 1, not 0 in standard mathematics in my
+>    experience.
+
+It is a usual approach in array languages and tensor calculus to define rank zero arrays/tensors
+to be isomorphic to scalars.  This becomes intuitive if you think about the product of the empty
+list, which is typically defined to be 1.  Also, if rank-zero array is always a trivial vector,
+for array (a : Ar [] (Ar [2,2] Nat)), what does `unnest a : Ar [2,2] Nat` should return?
+
+For example [this link](https://www.damtp.cam.ac.uk/user/tong/vc/vc7.pdf) says: A tensor of rank 0 is just a
+number, or scalar, T . Under a rotation, it doesn’t change: T 0 = T . A tensor of rank 1
+is a vector, while a tensor of rank 2 is a matrix.
+
+>  - 101: I can see why given the code things work out as they do, but this seems like an artifact of
+>    how the indexing is set up: the smallest representable array is 1 by 1, not 0 by 0... Presumably
+>    this is fine, because we do not need something like this, but is this really the standard
+>    indexing convention in machine learning?
+
+What do you mean by representable?  Any array that contains zero as one of its shape components
+is an empty array, and we have infinitely many of these.  Empty arrays do not contain elements
+(product of the shape is zero) but they are valid values, here is an example:
+```
+  x : Ar (2 ∷ 0 ∷ []) ℕ
+  x (i ∷ () ∷ [])  -- absurd pattern indicates that such index component couldn't possibly exist
+```
+They can be defined because the index space of an empty array is isomorphic to the empty type,
+and we can define a function from the empty type to any other type.  This is entirely standard
+treatment of arrays in array languages such as APL, SaC, etc.  I am pretty sure that NumPy
+has the same treatment of empty arrays.
+
+>  - 112: As a small side, zipWith should come from a more primitive operation which witnesses
+>    `Ar s (X × Y) ≅ Ar s X × Arr s Y` (so this is pointed cartesian functor), but this is neither
+>    this is a matter of aesthetics :)
+
+Sure, I agree, but we use `zipWith` in the code and do not really use the iso.
+
+>  - 130: this is another "X is just Y" of the highest order, but I suppose one can remark that `nest`
+>    and `unnest` ensure that `Ar - X` is a monoidal functor from `Shape` to `Type → Type` (with the
+>    latter monoidal product coming from composition). Hardly the a pressing issue, but I found it
+>    neat.
+
+Sure, happy to remark this.
+
+>  - 136: I think one should briefly explain what "pattern" does in Agda, since this sort of syntax is
+>    a bit more Agda-specific than what has come before. I think it can be guessed from context, but a
+>    footnote or just a description of what to search in the Agda docs might be helpful.
+>  - 218: "that 𝑞 is a point-wise addition of 𝑝 and 𝑞" should be r.
+
+Agreed, thanks for spotting this.
+
+>  - 228: Again, this footnote should be on the period.
+
+Oh, I learned something, thank you!  I was under the impression that footnotes go before punctuation, but I was wrong.
+
+>  - 221: Incidentally, why is this specialized to natural numbers? Surely this works fine for an
+>    arbitrary type A.
+
+Sure, then `a`, `b`, `c` has to become `List A`, but they are currently of type `S`.  It is just
+we only use this for shapes, so we didn't generalise it.
+
+>  - 465: I'm surprised that passing the environment as an instance argument is a safe idea here:
+>    could Agda not always update in the future to make this code just pass the empty environment
+>    instead?
+
+The type of environment is `Env Γ` where `Γ` is a parameter, whereas empty environment has the type
+`Env 𝜖`, so it wouldn't typecheck.  Secondly, the instance search wouldn't do such search, as `Env`
+is a function.  Finally, even if it were to find the empty environment that somehow magically fits,
+it would (potentially) contradict with the instance `𝜌`, and if more than one instance is found it
+reports a type error.
+
+>  - 716: "revers" to "reverse"
+
+Thanks.
+
+>  - 720: What is a "computational graph" in this context? Is it the dependency graph of expressions?
+
+Yes.
+
+>  - 725: 𝑤2= 𝑤1𝑤2 has a typo
+
+Yes, indeed, thank you.
+
+>  - Is the terminology adjoint standard? I've seen "partials" or "1-forms" for these objects.
+
+It seem to be pretty standard terminology that is also used on Wikipedia.
+
+### Reviewer D
+
+> Comments for authors
+> --------------------
+> - l50ff "follow [36]" here and everywhere: do not use citations as nouns
+
+Sure, happy to fix this.
+> 
+> - l99 Maybe add a footnote to the extend that `S ▷ P` is a container (I guess that motivated the letters `S` and `P`).
+>   However, I agree with your choice to not overload the paper with technical terms that do not help understanding.
+
+Happy to add a footnote.
+
+> - l103 It is worth remarking that an array of dimension 0 (shape `[]`) is a singleton ("skalar").
+
+We actually say this in line 86/87: "The `[]` shape describes an array of rank zero that contains exactly one element"
+
+> - l136 Is there a reason to use `ι n` instead of the standard `[ n ]` for the singleton `n`?
+
+Not really, so one could use `[ n ]` instead.
+
+> - l137ff What is called `sum` is really a `fold`, even if you only use it with addition later.
+
+Kind of, except it doesn't fold the unrolling of the array, it folds a hyperplane ar a time, so I am not sure whether the name fold helps here.
+
+> - l142 Why is the base case not simply `sum {s = []} f ε a = a []`?
+>   I tried this out with your Agda artifact and it works, one can simplify some proofs.
+>   An array of zero dimensions is a single scalar.  A combination of all its elements is thus just this one scalar.
+>   I think your starting point was the `foldr` intuition where the result always in corporates the neutral element,
+>   but maybe thinking in terms of non-empty folds (`fold1`) is more appropriate here.
+
+Exactly as you say, if `ε` is neutral to `f` then it doesn't matter.
+However, with lists, if you fold a singleteon list `fold f e [ x ]` you end up with `f x e` rather than `e`.
+As arrays of shape `[]` are singletons, I thought it is consistent to do the application of `f`.
+
+> - l169 I think the surprise goes away if the second hypothesis is written as $j ≤ n$ rather than $j + 1 < n$.
+
+My line of thought was: `i : Fin m` translates to `i < m` and `j : Fin (1 + n)` translates to `j < 1 + n` (as I explained in line 163).
+Then, for premises `a < b` and `c < d` it feels natural to conclude `a + c < b + d`, which is true, but it is also an overaproximation.
+I am happy to add the comment saying that `j < 1 + n` is the same as `j ≤ n`.
+
+> - l205 "fot" → for 
+> - l216 "that $q$ is a point-wise addition" → that $r$ is ...
+> - l220 "trenary" → ternary
+> - l232 "f (g x)" and "f (g x y)" are in the wrong font
+> - l246 "The implementation[s] of..."  Grammar and missing full stop.
+
+Sure, thanks for spotting those.
+
+> - l295ff A picture would help in the visualization of block formation (for low dimensions).
+
+We are a bit pressed on space, but I can try to add it.
+
+> - l299 "introducing blocked selections" insert `selp` here
+
+You mean `selb`, sure.
+
+> - l309 in the type you could have `Ar p (Ar s X)` instead of `P p → Ar s X` which would make it more symmetrical with `imapb`
+
+Sure, thanks.
+
+> - l419 why does `scaledown` take an `ℕ` and not an `ℝ`?
+
+For simplicity really.
+The definition of `E` keeps the type of reals completely opaque, e.g.
+it is a base type of the arrays of these shapes which you chose when you interpret the language.
+Any real type you chose should allow conversions from natural numbers.
+I could parametrise E with `ℝ`, or
+use `E Γ unit` as an argument, natural number seemed to be the simplest option.
+
+> - l456 "expressions in E Γ _is_" use parentheses "(E Γ _is_)"
+
+Sure, thanks.
+
+> - l696ff This section may offer opportunities for salvaging some space if you are struggling with the page limit:
+>   it feels things are spaced out much more generously than in the other sections.
+
+Yes, you are right, we can compress the list of equations.
+
+> - Section 5: There is a drop in quality in the exposition of this section.  It deserves some serious polishing.
+
+We'd be more than happy to try our best to improve this section.
+
+> - l716 "revers[e] mode"
+
+Sure.
+
+> - l717 Having to juggle with the aliases $y = w₃$ and $x = w₀$ obstructs the pattern.
+
+Fair enough.
+
+> - l720 What role is served by $f x y$ here?  Why not simply $z = \sin(xy + x)$.
+
+Agreed.
+
+> - l720 $sin$ should be $\sin$.
+
+Sure, thanks.
+
+> - l724 $w₂ = w₁ w₂$ I guess this should be $w₂ = w₀ w₁$.  The aliases even confuse you.
+> - l729 ∂y should be ∂z, should it?
+
+Yes to both, and thanks for noticing.
+> 
+> - l729 Please state what a "successor in the computational graph is", by example.
+>        E.g. for $w₂$ is $w₃$ the successor or are these $w₀$ and $w₁$?
+>        Make sure to include an example with more than one successor!
+> 
+> - l737ff I am totally lost here.  Please provide more calculation steps for each line.
+>          You got plenty of unused horizontal space here!
+> 
+> - l746 "If we inline all the $\bar{w}ᵢ$ definitions"  Again more steps please.
+>     If I do this myself, I am left with $\bar{w}₀ = \cos w₃ + (\cos w₃)²·x$ which isn't close to the results you report.
+
+I will redo the example providing more intermediate steps.
+
+> - l1000 This parenthesis never closes.
+
+This is a typo, thanks.
+
+> - l1015 So Futhark code is represented just by a `String`?
+>         Why don't you model the relevant parts by some abstract syntax?
+>         Can you even prove something about your normalization function when you juggle with monsters like `String → String` to represent a Futhark context?
+
+Yes, you are right, it would be nicer to prove safe translation to Futhark by
+defining a model of Futhark, and then relating the interpretation of futhark programs
+and evaluation of E.  This does require more work though, and in practice,
+code generation seemd to be the least interesting part of the work.
+We thought that the proposed way might be a reasonable compromise that gets
+us NBE without introducing much scaffolding.
+
+> 
+> 
+> 
+> Artifact: agda
+> ==============
+> 
+> Ar.agda
+> -------
+> 
+> I suppose some functions you define here can also be found in the standard library, e.g.
+> 
+> - `_++_ = Data.List.Relation.Unary.All.Properties.++⁺`
+> - `_∙_ = trans`
+> - `inject-left = _↑ˡ_`
+> - ...
+
+Sure, I recall that we had troubles with different versions of the standard library, so we just replicated some of the functions.  We can clean fix the library version and do a little cleanup.

### 2025-05-01 rebuttal/rebuttal.md
Tweak word
Finalise detailed comments.

@@
 
-As for correctness of AD, as we mentioned already: it takes about 20 lines
+As for correctness of AD, as we mentioned above: it takes about 20 lines
 of code to define all the AD rules (lines 979-830).  Most of these rules are
@@
 ### Reviewer A
+
+> l5: "strong correctness guarantees" -- like/namely?
+
+We make a list in lines 1197-1199, but we can surely put it here as well.
+
+> l80: "distinguishes it from most existing approaches" -- Sure, but there are also several rank polymorphic DSLs/libaries already, like Accelerate. Maybe mention them somewhere?
+
+Accelerate is mentioned 1168-1169, but we can mention it here as well.
+Also, a small subtlety, Accelerate allows only for static ranks as far as I am aware, where as the proposed array theory supports dynamic ranks as well.
+
+> l102: say somewhere that S is simply a list of natural numbers
+
+We say this in line 86.
+
+> l102: It would be helpful to say here what the intended semantics of Ar [n1, ..., nk] is, i.e. R^{n1 x ... x nk} = R^{n1} (x) ... (x) R^{nk}.
+> 
+> l111: maybe note that you can define n-ary zipWith's even.
+> 
+> l119: corresponds
+
+Sure, thanks.
+
+> l136: why call this a sum rather than a fold? this is particularly confusing as later sum is just a sum.
+
+My thought process was as follows.  This is not quite fold because it does not travers the unrolling, instead it recurses through the shape.  However, maybe sum is not the best name, happy to change.
+
+> l136: "pattern" -- quickly recall pattern syntax for readers less familiar with Agda
+> 
+> l160: "Dec" -- Just give the definition! It is not that hard and your current description is not enough to understand what it does.
+
+Sure, will do.
+
+> l160: "\exists" -- Explain. Contrast with "\Sigma". Does the choice matter here?
+
+They are the same, it is just a shorter syntax.  I am happy to explain this.
+
+> l172-l180:  This is currently super unclear and should probably be rewritten.
+
+Ok, I'll have another go at this.
+
+> l204: for
+
+Sure, thanks.
+
+> l204-205: It is unclear to me what the alternative is.
+
+The alternative is to move `sum` inside the `imap` and compute the tile in the input array.
+
+> l220: ternary
+> 
+> l224: Explain instance argument syntax {{}} here.
+> 
+> l220-225: Say in prose what these relations do/capture.
+> 
+> l235: Remind readers of Agda's notation for identity types.
+
+Will do, thanks.
+
+> l258: Equations relating slide and backslide?
+
+Good idea, thanks!
+
+> l297: "the local neighbourhood"
+> 
+> l298: rephrase sentence. grammar does not work.
+> 
+> l299: blocked selections "selb"
+
+Thanks, will do.
+
+> l329-342: very nice
+
+Thanks :)
+
+> l375: "use non-trivial dependencies within constructors" -- Please explain.
+
+Will do.  The point is that constructors are polymorphic in shapes and proofs in `imapb`, `selb` depend on these shapes.
+
+> l401: "imaps" -- isn't this normally called build or generate. Why stray from that? 
+> l402:  "sels" -- isn't this just indexing? why not call it that?
+
+I am not sure about build/generate.
+The intuition here is that constructor is an index map (maps the value per index) and destructor `sel`ects a scalar therefore `sels`.
+I do not have a strong opinion about these names.
+> 
+> l401-402: say in words what imap(s/b) and sel(s/b) are supposed to do.
+> 
+> l411: explain how zero-but gives a conditional.
+> 
+> l429: that we call imaps 
+> 
+> l433: equality.
+> 
+> l494: verb missing
+
+Will do.
+
+> l501-552: isn't this all standard? is the only point to introduce your notation?
+
+It is standard.
+However, I was criticised before that without these definitions it is hard to understand some of the expressions that use weakening, `skip`, etc.
+So I thought it would be a good idea to show basic definitions.
+
+> l554-694: Interesting! I didn't know this technique. Is this a novel contribution? If so, maybe list it in the intro?
+
+The basic idea is old, but I think that this hasn't been done for an intrinsically-typed language.  Happy to add this.
+
+> l702-735: Why so much white space? This seems like a perfect place to insert some diagrams/graphs!
+> 
+> l714-718: "inside-out....outside-in" -- This is quite vague and does not help the reader understand the difference between forward and reverse mode.
+> 
+> l716: reverse mode
+> 
+> l719: we can compute partial derivatives with respect to all of the inputs
+> 
+> l729: "computational graph" -- what is that? it comes out of nowhere here!
+
+You are right, I'd like to redo the beginning of section 5.
+
+> l786-791: please explain a bit more here.
+
+Sure.
+
+> l797-830: What is the complexity of the resulting algorithm? Is it as expected from reverse AD?
+
+I answered this above.
+
+> l925: is this how you fix expensive one-hot arrays?
+
+It is mainly 935-937 that does this.
+We took a compromise here by keeping the language simple.
+As we need to optimise AD-generated code anyway, we can also try to resolve this problem through rewrite.
+
+> l942-944: :-(
+
+This is surely not ideal, but this is pretty much inevitable state in optimising compilers.
+Specifically when languages are high-level, the quality of generated code crucially depends on the set of optimisations and even the order of running them.
+Given that people make research careers on applying machine learning to figure the best optimisation strategy, this is still unsolved problem.
+
+> l1079: What about even larger datasets?
+
+The structure of the code is such that there is a fixed-size kernel that runs backpropagation, and the size of input only indicates how many times this kernel will be launched (sequentially).
+Therefore, it is very unlikely that bigger input sizes would change the behaviour.
+However, you are right, we are happy to run a bigger dataset and report the results.
+
+> l1122: You may want to include some comparisons here to
+> de Vilhena, Paulo Emílio, and François Pottier. "Verifying an Effect-Handler-Based Define-By-Run Reverse-Mode AD Library." Logical Methods in Computer Science 19 (2023).
+> Paszke, Adam, et al. "Getting to the point: index sets and parallelism-preserving autodiff for pointful array programming." Proceedings of the ACM on Programming Languages 5.ICFP (2021): 1-29.
+> Smeding, Tom J., and Matthijs IL Vákár. "Efficient CHAD." Proceedings of the ACM on Programming Languages 8.POPL (2024): 1060-1088.
+> Smeding, Tom, and Matthijs Vákár. "Parallel dual-numbers reverse ad." arXiv preprint arXiv:2207.03418 (2022).
+
+Thanks, will do.
+
+> l1145: correctness 
+> 
+> l1171: "Even though our support for.."
+
+Thanks.
+
+> l1185-1187: Explain more please. I don't understand the difference.
+
+Sure, happy to explain more.
+In a nutshell, shape annotations in both SaC and Futhark are somewhat like types in Python, you can use them if you want to, but they could be ignored and do not provide guarantees.
+Dynamic rank polymorphism means that array ranks are only known at runtime.
+
+> l1197: "certain functions being inverses" -- where do you use this?
+
+Line 160 and 242 mainly to ensure that subtraction is an inverse to addition.
 

### 2025-05-02 rebuttal/rebuttal.md
Edit rebuttal intro.

@@
 polymorphism, to transformation via AD, to optimisation and code generation. Not
-all parts of the whole are equally completely specified: while rank polymorphism
-is the most developed and fully verified part, the AD is performed within a
-well-shaped and well-scoped DSL, verifying absence of bounds and scoping errors,
-transformations that we run prior code generation guarantee semantics preservation
-and code generation while implementing NBE does indeed use aptly termed
+all parts of the whole are equally completely specified. Rank polymorphic
+array theory
+is the most developed and fully verified part. The AD is performed within a
+well-shaped and well-scoped DSL, verifying its termination and absence of bounds and scoping errors.
+Transformations that we run prior code generation guarantee semantics preservation.
+Code generation implements NBE but does indeed use aptly termed
 `String → String` monster.
 
-In an ideal world, all parts would of course come with formal specifications
+In an ideal world, all parts would of course come with full formal specifications
 and accompanying proof, but in practice some parts remain an entirely separate
-(and nontrivial) research effort to verify functionally correct (such as AD),
-while others can be handled using known techniques (such as code generation),
-but are somewhat time consuming. We wish to show that even in those cases where
+(and nontrivial) research effort to verify for functional correctness (such as AD).
+Other parts can be handled using known techniques (such as code generation),
+but are somewhat time consuming. Despite all this, we wish to show that even in those cases where
 resources preclude full verification, Agda can still be used as a practical

### 2025-05-02 rebuttal/rebuttal.md
Fixes.

@@
 
-The key driver of the technique is the `Prefix` relation that checks whether
-the context is a prefix of another and this is done automatically using 
-the instance search.  This bit is solid, and for the DSLs with de Bruijn
-variables, this works very well.  You'd have to define wrappers for all
-the binders within your DSL, but this is straight-forward.
+The key driver for the technique is the `Prefix` relation that automates checking whether
+one context is a prefix of another one using 
+instance search.  This bit is solid, and for intrinsically-typed DSLs with de Bruijn
+variables, this works very well.  One has to define wrappers for all
+the binders within the DSL, but this is straight-forward.
 
-The limitation of all techniques that are based on instance search is
-a bad complexity of instant search implementation.  When many things
-turn into instances, instance search may have to traverse very large
+The limitation of all techniques that are based on instance search has
+to do with bad complexity of instant search implementation.  When many things
+become instances, instance search may have to traverse very large
 search spaces (e.g. evaluating a large data structure of a constant
-size, and so on).  However, this is not a problem related to DSL wrappers,
-as we rarely define a very large number of local variables, we can
-typically split a big expression into subexpressions and check them locally,
-as we do with primitives in our CNN definition.
+size, and so on).  However, this is a general concern rather than a
+practical problem related to DSL wrappers.
 

### 2025-05-02 rebuttal/rebuttal.md
Fixes

@@
 
-There is quite a significant difference, as this work
-creates the actual data structure (`class Hoas`) that specifies construct
+There is quite a significant difference, as the quoted work
+creates the actual data structure (`class Hoas`) that specifies constructs
 of the untyped language syntax.  Then untyped terms are scope- and type-checked
-using instance resolution mechanism, and if all the instances match, they are
-turned into unscoped de Bruijn, scoped de Bruijn, and simply-typed de Bruijn
-representations.  Our technique never defines a particular HOAS syntax.
-Every time we use our wrappers we conceptually write the term in `E`,
+using instance resolution mechanism, and if all the instances are found, the expression is
+turned into unscoped de Bruijn, scoped de Bruijn, or simply-typed de Bruijn
+representation.  Our technique never defines a particular HOAS syntax.
+Every time we use our wrappers, we conceptually write the term in `E`,
 our wrappers simply help us to automate computation of well-typed
@@
 `selb`, etc depend on the shapes bound by the given constructor.  It is unclear
-to me whether the proposed technique can easily handle such a case.  While this
-might be the case, it does not look to very simple.  Once again, conceptually,
-all our technique does, it automates computation of the offsets within the context.
+whether the proposed technique can easily handle such cases.  While this
+might be the case, the technique does not look very simple to use.  So we
+took a conceptually simpler approach, which just automates computation
+of offsets within the context.
 

### 2025-05-02 rebuttal/rebuttal.md
Minor tweaks.

@@
 We thank the reviewers for their comments. Before responding to specific
-questions below, we wish to explain the intent of our paper in a way that we
-perhaps failed to do in the paper itself.
+questions below (which we do not expect reviewers to read fully), we wish to
+explain the intent of our paper in a way that we perhaps failed to do in the
+paper itself.
 
@@
 TensorFlow, but to show that the performance of our system seems sufficient to
-be useful. More complicated benchmark applications, including scientific
-programs outside of machine learning, remain future work.
+be useful, and that inefficiencies are not likely due to overheads imposed by
+formal verification. More complicated benchmark applications, including
+scientific programs outside of machine learning, remain future work.
 

### 2025-05-02 rebuttal/rebuttal.md
Minor
More fixes.

@@
 
-As mentioned above there are two sources of inefficiency: excessive copying and
+As mentioned above, there are two sources of inefficiency: excessive copying and
 `zero-but` representation.  Excessive copying is a known problem in array languages,
@@
 
-My thought process was as follows.  This is not quite fold because it does not travers the unrolling, instead it recurses through the shape.  However, maybe sum is not the best name, happy to change.
+My thought process was as follows.
+This is not quite `fold` because it does not traverse the unrolling, instead it recurses through the shape (hyperplanes).
+However, maybe sum is not the best name, happy to change.
 
@@
 I am not sure about build/generate.
-The intuition here is that constructor is an index map (maps the value per index) and destructor `sel`ects a scalar therefore `sels`.
-I do not have a strong opinion about these names.
+The intuition here is that the constructor is an index map (maps the value per index) and the destructor `sel`ects a scalar, hence `sels`.
+I do not have a strong opinion about these names and happy to change them.
 > 
@@
 It is standard.
-However, I was criticised before that without these definitions it is hard to understand some of the expressions that use weakening, `skip`, etc.
+However, I was criticised before that when these definitions are omitted it is hard to understand some of the expressions that use weakening, `skip`, etc.
 So I thought it would be a good idea to show basic definitions.
@@
 
-The basic idea is old, but I think that this hasn't been done for an intrinsically-typed language.  Happy to add this.
+I have seen the basic idea before, but I do not think that this has been applied for an intrinsically-typed language.  Happy to add this to the introduction.
 
@@
 Specifically when languages are high-level, the quality of generated code crucially depends on the set of optimisations and even the order of running them.
-Given that people make research careers on applying machine learning to figure the best optimisation strategy, this is still unsolved problem.
+Given that people make research careers on applying machine learning to figure the best optimisation strategy, this is still an unsolved problem.
 
@@
 Therefore, it is very unlikely that bigger input sizes would change the behaviour.
-However, you are right, we are happy to run a bigger dataset and report the results.
+However, we are happy to run a bigger dataset and report the results.
 
@@
 list, which is typically defined to be 1.  Also, if rank-zero array is always a trivial vector,
-for array (a : Ar [] (Ar [2,2] Nat)), what does `unnest a : Ar [2,2] Nat` should return?
+for array (a : Ar [] (Ar [2,2] X)), how could `unnest a : Ar [2,2] X` ever produce a result?
 
@@
 is an empty array, and we have infinitely many of these.  Empty arrays do not contain elements
-(product of the shape is zero) but they are valid values, here is an example:
+(product of the shape is zero), but they are valid values, here is an example:
 ```
   x : Ar (2 ∷ 0 ∷ []) ℕ
-  x (i ∷ () ∷ [])  -- absurd pattern indicates that such index component couldn't possibly exist
+  x (i ∷ () ∷ [])  -- absurd pattern indicates that such an index component couldn't possibly exist
 ```
@@
 
-Sure, I agree, but we use `zipWith` in the code and do not really use the iso.
+Sure, I agree, but we use `zipWith` in the code and we do not really use the iso.
 
@@
 
-Sure, then `a`, `b`, `c` has to become `List A`, but they are currently of type `S`.  It is just
-we only use this for shapes, so we didn't generalise it.
+Sure, then `a`, `b`, `c` has to become `List A`, but they are currently of type `S`.
+We only use these relations for shapes, so we didn't generalise it.
 
@@
 
-The type of environment is `Env Γ` where `Γ` is a parameter, whereas empty environment has the type
-`Env 𝜖`, so it wouldn't typecheck.  Secondly, the instance search wouldn't do such search, as `Env`
+The type of environment is `Env Γ` where `Γ` is a parameter, whereas the empty environment has the type
+`Env 𝜖`, so such an instance wouldn't typecheck.  Secondly, the instance search wouldn't do such search, as `Env`
 is a function.  Finally, even if it were to find the empty environment that somehow magically fits,
-it would (potentially) contradict with the instance `𝜌`, and if more than one instance is found it
+it would (potentially) clash with the instance `𝜌`, and if more than one instance is found it
 reports a type error.
@@
 
-Kind of, except it doesn't fold the unrolling of the array, it folds a hyperplane ar a time, so I am not sure whether the name fold helps here.
+Kind of, except it doesn't fold the unrolling of the array, it folds a hyperplane ar a time, so I am not sure whether the name fold helps here. I am not opposed to changing the name though.
 
@@
 Exactly as you say, if `ε` is neutral to `f` then it doesn't matter.
-However, with lists, if you fold a singleteon list `fold f e [ x ]` you end up with `f x e` rather than `e`.
-As arrays of shape `[]` are singletons, I thought it is consistent to do the application of `f`.
+However, with lists, if you fold a singleton list `fold f e [ x ]` you end up with `f x e` rather than `e`.  As arrays of shape `[]` are singletons, it seems reasonable to do the application of `f`.
 
@@
 My line of thought was: `i : Fin m` translates to `i < m` and `j : Fin (1 + n)` translates to `j < 1 + n` (as I explained in line 163).
-Then, for premises `a < b` and `c < d` it feels natural to conclude `a + c < b + d`, which is true, but it is also an overaproximation.
+Then, for premises `a < b` and `c < d` it feels natural to conclude `a + c < b + d`, which is true, but it is also an overapproximation.
 I am happy to add the comment saying that `j < 1 + n` is the same as `j ≤ n`.
@@
 
-We are a bit pressed on space, but I can try to add it.
+We are a bit pressed on space, but I can try to add this.
 
@@
 and evaluation of E.  This does require more work though, and in practice,
-code generation seemd to be the least interesting part of the work.
+code generation seemed to be the least interesting part of the work.
 We thought that the proposed way might be a reasonable compromise that gets
@@
 
-Sure, I recall that we had troubles with different versions of the standard library, so we just replicated some of the functions.  We can clean fix the library version and do a little cleanup.
+Sure, I recall that we had troubles when collaborators used different versions of the standard library, so we just replicated some of the functions.
+We can clean fix the library version and do a little cleanup.

### 2025-05-06 rebuttal/response-to-E.md
Response to the new reviewer.

This is a reply to the reviewer E which is written as a part of rebuttal.
As the review came after the rebuttal period was over we are writing a
separate response.

Thank you for your excitement and very in-depth reviews, we found them
tremendously helpful.

Before answering your questions, I'd like
to make a general remark that relate to a number of your comments.
I suppose that you are a kind of reader who approaches
problems through categorical, type-theoretical or other foundational constructions.
This is great, I also like to do this.  Unfortunately, I have been criticised too
many times for this.  I was told that instead of getting straight to the point,
I introduce too many abstractions that are not necessary.
Therefore, the mindset I had when writing this paper was: expose just
enough constructions so that the key ideas are well-defined and non-ambiguous.
I might have gone too far with this, and I'd be more than happy to introduce
more abstract notions into new revision, but finding a good balance is difficult.

> Review #55E
> ===========================================================================
> 
> Overall merit
> -------------
> A. Good paper, I will champion it
> 
> Reviewer expertise
> ------------------
> 4. Expert
> 
> Reviewer confidence
> -------------------
> 3. High
> 
> Paper summary
> -------------
> This paper displays a synthesis of multiple techniques in 'modern' FP,
> targeted at the specimen application/running example of the MNIST
> benchmark in machine learning. It showcases the use of advanced type
> systems in FP, in this case, that of the Agda implementation of type
> theory, as well as that of the Futhark language for high-performance
> computation.
> 
> Specifically, the authors
> 
> * develop a rank- and bounds- safe language of rank-polymorphic tensor
>   algebra, including sums and convolutions necessary for purely
>   functional implementations of convolutional neural networks (CNN;
>   such as those required in the MNIST example)
> 
> * develop a scope-, type-, and rank- safe embedded *differentiable*
>   DSL for writing CNNs, paying careful attention to issues of
>   differentiability, sharing of subcomputations (eg. induced by
>   `let`), and impedance matching against the eventual chosen Futhark
>   backend for program extraction; together with the rudiments of a
>   semantically-justified theory of optimisation of DSL programs;
>   by analogy with LINQ, this is 'language-integrated gradient' computation?
> 
> * exhibit extraction to Futhark, with some specimen performance
>   comparison of the extracted code against the SOTA PyTorch and
>   TensorFlow frameworks
> 
> Comments for authors
> --------------------
> I really liked this paper... even though I think there are (very) many
> things that I think you could, and probably should, do to improve it
> if it ends up being accepted. The summary above captures the broad
> picture of what your paper achieves, and that alone is impressive.
> 
> I particularly appreciate how the paper demonstrates how advanced
> techniques in FP really do seem now to be available for end-use
> applications such as implementing CNNs to run on the MNIST example
> problem, and the end-to-end from Agda through to Futhark seems very
> convincing, despite each being very niche languages.
> 
> Detailed comments
> 
> p.1
> This perhaps does mean you can get away with some of the more
> rhetorical flourishes in the Introduction, but I think that the title
> is a bit misleading ("Correctness meets Performance: from Agda to
> Futhark" seems as though it would be a slightly more honest version?
> the second paragraph of the introduction makes that pipeline/trade-off
> more explicit), and the Abstract reads a bit flat in the context of
> the more dramatic Intro.

I agree with the newly proposed title, and I am happy to imorove abstract.

> p.2
> The 'running example' of the SAC implementation of a CNN in fact only
> appears eventually on p.7, and it might be better to at least show the
> original code (or fragments) before diving into Agda; few readers will
> be expert in both, possibly neither, so some handholding would help,
> not least to prime the reader to accept that the eventual Agda
> implementation is indeed faithful to the original.

My reasoning here was that (a) SaC is quite an esoteric language for the
ICFP audience, and (b) the full code listing is relatively big, so I
decided to skip the SaC code entirely.  I agree with you about handholding,
and I am happy to summarise the code in SaC before diving into Agda.

> You introduce the fundamental representation of higher-rank tensors
> (arrays) as a container type, in terms of 'shapes' and 'positions',
> without any mention of the pioneering work in this area by Abbott et
> al. ("Categories of Containers", 2003), or that of Joyal et al. at
> UQAM on combinatorial species.

I should definitely mention this.  I didn't want to introduce too much
theoretical constructions here, but I will definitely reference this work
that I am perfectly aware of.

> Similarly, you launch into the discussion of the Agda representation
> without taking the small amount of space to explain that a
> 1-dimensional array (vector) $X^n$ can be represented in terms of the
> shape (dimension) `n : ℕ`, positions (indices) `i : Fin n` (thereby
> ensuring bounds safety for indexing operations), as a function `Fin n
> → X`, and that your `Ar` construction then generalises this to
> higher-rank via *lists* of such dimensions. For the non-Agda
> specialist, this little bit of hand-holding might go a long way!

Good idea, thanks, will do.

> p.3
> You do a lot of heavy lifting here for the non-specialist, while not
> really discussing trade-offs in the details of the representation:
> 
> * why define the Shape type `S` directly as a `data` type definition,
>   and not simply in terms of Agda's `List` type constructor? (more
>   generally: what/how much could you simplify your implementation by
>   using off-the-shelf constructions in Agda's standard library?)

My reasoning here was as follows:  these definitions are fundamental to the
story, and they can be specified very shortly, so I thought I would write them
directly which avoids assumptions about familiarity with Agda's standard library.
Otherwise, you are absolutely right, I am working with a 
`List Nat <| All Fin` container.  I do mention that S is a list of Nat
in line 86, I can stress this harder.

As for simplifications from using List from the standard library, I wouldn't
say that too many things become simpler.  You can see that our `Ar` module is
about 250 lines of code.  A small part of it could be replaced with constructions
from stdlib, but this is not a significant gain.

> * why define the Positions type family `P` directly as a `data` type
>   definition, rather than as a definition by recursion on `S`, as an
>   iterated product type?

You can pattern-match on indices directly, which sometimes look nicer,
but I agree that there isn't much difference.

> * given that you define `Ar s X = P s → X`, presumably to take
>   advantage in the back-end of arrays as 'purely functional', why do
>   you not discuss the trade-offs (esp. wrt elimination of those
>   iterated products defining `P`, including the troublesome case `s =
>   []`, introducing a redundant unit type representation) between this
>   version and a direct *recursive* implementation such as
>   
>   Ar : S → Set → Set
>   Ar [] X = X
>   Ar (n ∷ s) X = Fin n → Arr s X
>   
>   which would have the advantage(s) of:
>   
>   - avoiding having to distinguish scalars, at shape `[]`, from
>     functions from `P []` to scalars; similarly in the 1-dimensional
>     (vector) case
>   
>   - avoiding the marshalling/unmarshalling of iterated products of
>     positions via your `nest`/`unnest` operations, in favour of
>     iterated function composition
> 
>   - avoiding most, if not all, of the explicit operations on
>     positions, `_⊗ₚ_`, `split` etc. in favour of direct implementation
>     on the `Ar s X` types themselves (more on this later)

This is a good point, and I should add this discussion.  However, I would
like to note that the mentioned benefits are only acheived when working
with statically-known shapes.  Most of the interesting combinators are
defined rank-polymorphically, e.g quantifying over `s` and `p` shapes.
In this case, we still need nest/unnest (or transport over the equality)
and split-like operations for the shapes `s ++ p`.  Also, with this 
definition one needs to introduce selection separately, where now
it is just function application.

> l.109 Rather than merely consider `Ar s` for fixed `s` to be an
> Applicative functor, wouldn't it be better to say that `Ar` is a
> *graded* applicative, graded with respect to the monoid structure on
> ranks-as-lists of dimensions? Not only that, but the `nest`/`unnest`
> isomorphism exhibit a corresponding 'linear'/'monoidal closed' structure (scare
> quotes because I'm not even sure what the correct terminology is here
> for graded functors)

This confusion that you have was exactly the motivation to avoid this
discussion entirely.  If we fix element types then I would call `Ar - X`
a monoidal functor.  I am happy to mention this.

> Similarly to the above,
> 
> * why go the trouble of defining 1-dimensional versions of `sum`
>   etc. and end up with definitions which are mismatched/off-by-one
>   between the 1-dimensional and general cases? Eg. the definitions of
>   `sum₁ : (A → A → A) → A → Vec n A → A` and `sum : (A → A → A) → A →
>   Ar s A → A` do *not* agree in the case `s = ι n`, ie. `sum₁ {n = n}
>   f e` and `sum {s = ι n} f e` are *not* extensionally equal unless
>   `f e` is the identity function. Did you intend this? If so, please
>   explain why.

Originally, the reasoning was that sum1 is just a helper function, and as
sum folds over dimension, we typically want `fold f [x] e` to be `f e x`.
However, for the purposes of this work, we are interested in the cases
where `e` is neutral to `f`, so indeed we can simplify this and make it
less confusing.

> p.4
> I don't see the need for *decidability* of the `_⊖_` operation (and
> those which build on it later, etc.), rather than simply returning a
> `Maybe` option type, as the `slide` function (and subsequent friends)
> are going to use the `nothing` case to return a default, so full
> decidability, while impressive, seems like overkill.

Actually, I tend to disagree.  The trouble is that Maybe (specifically
in this case) can actually introduce off-by-one errors.  One can return
`nothing` in the case where the value actually exists.  This did happen
in one of the implementations in this particular function.  Therefore,
while it is a bit of an overkill, this is a useful safety measure.
I am happy to motivate this a little bit more in the text.

> Similarly, the discussion of the type isomorphism seems muddled, when
> what is really going seems to be that are going to some lengths to
> skilfully avoid having to negotiate the equality `m + suc n ≡ suc (m +
> n)` in the type of positions?

It is a bit more subtle, a for `i : Fin m`, `j : Fin (1 + n)` their sum
(when using`+` from Fin's) is `i + j : Fin (toℕ i + suc n)`, but we need
`Fin (m + n)`.  As for the discussion about isomorphism, this is literally
a reply to the one of the reviewers comment from the previous submission
attempt who said: your definition of ⊕ is overly complicated, surely
you can simplify this through the isomorphism...   I am happy to have another
go at explaining this bit.

> p.5
> line 202 Should this be `sum` or `sum₁` in the definition of `conv₁` (see also above)?

I would define them to be equivalent in this case this would be a moot
point.

> line 205 Typo: "for" in place of "fot"

Sure, thanks!

> Section 3.2
> 
> This seems needlessly over-elaborate, at the potential cost of losing
> readers who are not Agda experts:
> 
> * the use of `instance`s, and instance resolution, for the sake of
>   optimising certain definitions seems a marginal gain, at the cost of
>   a clear exposition (and is really an Agda-dependent implementation
>   artefact, rather than part of your story about representation within
>   a dependently-typed language)

Well, one benefit is that I do not have to construct any pointwise proofs
in the definition of forward.  My thinking was: I am going to use instances
anyway to do syntactical wrappers for the DSL, so maybe I could make the
definitions nicer.  I am happy to take another go and see whether removing
instances adds a lot of proof obligations.

> * the use of the 3-place pointwise relation type former makes sense,
>   although as it is only used in two cases, I would be tempted simply
>   to inline one use (for addition), and indicate that multiplication
>   is similar
> 
> * similarly, as far as I could tell, you do not at any point exploit
>   the full generality of the `cons` constructor in maintaining the
>   invariant `R m n k`, when in the two uses you make, these could be
>   inlined simply with `k = m + n`/`k = m * n` *definitionally*

I arrived at this definition in attempt to keep my presentation compact.
However, I am happy to attempt inlining.

> * moreover: the one use of the 2-place pointwise construction, for the
>   case of the `suc` function, completely obscures, via the relational
>   version `suc p ≈ q`, that in all such uses, any such `q` is
>   *definitionally* equal to `sucₛ p = List.map suc p`, and as such, would
>   considerably simplify the types of functions which make use of a
>   combination of this and the `p + q ≈ r` invariant (see below)

`suc` is kept as equation on purpose, because I literally translate
these combinators into constructors of E.  It turns out that it is
helpful to keep as little computations in type indices as possible.
This is not so relevant for the definitions within Agda, but I thought
this consistency would make it easier for the readers to grasp constructors
of E.

> p.6
> The "little plumbing" required for rank-safe 'subtraction' is
> considerably simplified if you only require a `Maybe` option type to
> be returned (see above)!

I agree, but I would use the same argument here --- Maybe does not prevent
me from sayin `nothing` where there is actually something.

> The types of `slide` and `backslide` I found to be very difficult to
> read, combining as they do, in two different argument orders, appeal
> to both the `p + q ≈ r` and `suc p ≈ q` invariants. Moreover, here, as
> later with `conv` and `mconv`, there is an explicit mention of the
> function space arrow `P s → ...` when it seems as though this should
> simply be folded into the type, given the central `Ar s X`
> abstraction, namely to give the following types to the operations:
> 
> slide     : p +ₛ q ≈ r → Ar r A → Ar p (Ar (sucₛ q) A)
> backslide : p +ₛ q ≈ r → A → Ar (sucₛ q) A → Ar p (Ar r A)
> conv      : p +ₛ q ≈ r → Arr r R → Arr p R → Arr (sucₛ q) R
> mconv     : p +ₛ q ≈ r → Arr r R → Arr s R → Arr s (Arr p R) → Arr s (Arr (sucₛ q) R)
> 
> (systematically identifying `Arr s (Arr p X)` with `Arr (s ⊗ p) X` via
> the `nest`/`unnest` isomorphisms)

Your definitions do look nice, but as I was saying, I was trying to keep
computations out of array indices.  I'd have to think whether there is
a good way to introduce something like `sucₛ` that expands to a point-wise
equation and the array.

> NB. the `conv` and `mconv` constructions are moreover generic in any
> particular (semi)ring structure on `R`, so that `conv₁` can (should?)
> be seen in terms of the semiring instance for `R = ℕ` etc.

Yes, this is true, I am happy to mention this, but I didn't want to add
extra levle of abstraction by parametrising the entire construction by a
semiring.  I can certainly mention this in the text or in the footnote.

> p.7
> The "running example" finally makes an appearance, but without ever
> having been explicitly mentioned (nor the role that the `forward`
> function plays in its development/implementation)... some more words
> earlier, and here, would help make that bridge.

Sure, happy to expand.

> The definition of `forward` is given in terms of Agda's `let`
> construct, when it might be smoother simply to give a definition using
> `where` clauses, not least because Agda's `let` actually gives rise to
> inlined substitution in the typechecked code, and thus potentially a
> loss of sharing upfront.

I agree in principle, but note that we do not really care about performance
of this shallowly-embedded definition.  One reason why I used let instead
of where is because it rhymes better with the definition of `cnn` in line
670.

> Section 4 The embedded DSL
> 
> p.8
> 
> While this now might be regarded as 'standard', or 'folklore', the use
> of dependent types to ensure type- and scope- safe embedding of DSLs
> into Agda ought to have associated citation(s); anticipating 4.3, the
> ICFP/JFP papers of Allais et al. could/should be cited here.

Sure, happy to add this.

> The introduction of `unit` as a synonym for shape `[]` seems
> excessive, and pointless, here, compared to other approaches sketched
> above for dealing with scalar types as an instance of the fundamental
> `Ar` type.

While I agree with your observation that Ar [] X can be defined as X in
case of shallow embedding, it seems that in case of deep embedding it does
not seem that elegant.  Note that the types used in E distinguish between
arrays and indices, yet they do not have a notion
of a base type (scalar).  There is some beauty to it: all we have is
`ar s` and `ix s` where `s` is a shape.  We can define rank-polymorphic
binary operations that work on arrays of the same shape (including scalars).\
The same goes for `sum`.
If we were to add more distinct cases in the definition of `IS`, we'd probably
have to add even more constructors maybe fix a type of reals (or add add it
as a parameter), or index with something larger than `IS`.  Given that we
want to define generic constructs in E that abstract over shapes `s` and `p`
it is not obvious to me which type encoding is the best.

> Similarly, why postpone the introduction of syntax for `bin plus` and
> `bin mul` to p.9 ll.423--4, when they would make more sense to be
> introduced when the `Bop` type is introduced? Better still, why waste
> a `pattern` synonym, when you could write smart constructors to build
> in optimisation for the cases of addition of 0, or multiplication by
> 1, which you both allude to (indirectly) later, and which make use of
> these units being part of the *syntax* of the language (and hence
> matchable on, IIUC?)

I agree, we used to have more primitive operations, but we got rid of them
for the sake of brevity.  We can surely inline these now.

> p.9
> 
> The definition of `data E` would surely look better (at the cost of
> some space, perhaps?) as a Figure, rather than inlined?

Sure, I can do this.

> Similar remarks apply here to the types of the DSL analogues of
> `slide`, `backslide` etc. wrt the use of the complex shape invariants

The same answer as before applies :)

> It seems a shame that you can't/don't permit a type structure for `E`
> which supports the `nest`/`unnest` isomorphisms. Perhaps this is now an
> artefact of extraction towards Futhark, but if so, this merits a bit
> more discussion, esp. given that you subsequently consider
> 'normalisation' of `E` programs before extraction, so that you might
> (and the reader, this one included) consider a richer DSL type
> structure, with restrictions to 'Futhark-conformant' types a later
> phase in the extraction pipeline.

I'd be happy to discuss this more in the text, but I am not sure why
do you think
we need to have both nested arrays and arrays of product shape, given
that we have established that they are in isomorphism.  The language
chooses a single representation (using product shape) and then we
have `imap` that performs what in shallow embedding would be acheived
with a lambda over `nest` applied to some array.  I thought that
keeping a representation small is a good thing.  As for Futhark,
it doesn't really matter whether we convert a homogeneously nested array,
or an array of a product shape, they will have the same representation
in Futhark.  The only thing that Futhark requires is static ranks,
so that we could generate a loop-nest-like construction.

> p.10
> 
> The specifics of the axiomatisation of a `Real` type, with suitable
> primitives, and their derivatives, seems something which doesn't
> require this level of detail; or rather, it's a level of detail which
> could be more economically explained in terms of standard library
> constructs such as `Ring` or `Semiring`, extended with `logistic`/RelU
> as an abstract operation... together with its derivative, rather than
> having to litigate the sharing problems associated with its concrete
> definition?

Sure, but doesn't relying on stdlib excludes the readers that are not familiar
with it?  That was my motivation to inline the definitions.  I am happy
to include generalised explanation into the text.

> In particular, the choice of a unary minus, without a companion
> definition of binary subtraction, seems to generate unnecessary
> clutter downstream. Similarly, an explicit `fromℕ` operation, used
> only at values 0 and 1, rather than the (obvious) additive and
> multiplicative units from an ambient `(Semi)Ring` structure, etc.

I agree with the general comment.  My idea was to define the smallest
set of operations on Reals that suffice to give an interpretation of E.
As for unary minus, it is just a way to introduce subtraction, as
`a - b` = `a + (-b)`.  The fromℕ is also used in `scaledown` which we
actually use a few times in the resulting definition of the cnn.  I am
happy to explain these in the language of Rings/Fields.

> p.11--14, Sections 4.2, 4.3
> 
> This seems to be an extended exposition of 'standard' (cf. Allais et
> al. above) constructions associated with de Bruijn representations of
> binding, but dressed up so that the nature of the Kripke function
> space semantics for binding forms is made more obscure than necessary:
> the Kripke idea of quantifying over context extensions in order to
> give semantics to object-level functions/`let`-bindings in terms of
> meta-level functions is at the heart of the Allais et
> al. papers... and goes much further back in the literature on the
> metamathematics of (representations of) binding. I would dearly love
> to see in particular 4.3 rewritten so as to make these ideas easier to
> grasp for the non-expert. As they stand, they're fine but
> incomprehensible (sic).

I am happy to have another go at this section and refer to Kripke.
Perhaps I was a bit overprotective while writing the prose as I am often
criticised for introducing too much theoretical constructions that can
be avoided.  For these particular sections, I was simply trying to
show enough definitions so that further constructions are not ambigous.
I will try to find a good balance in the new description.

> p.14
> The definition of `sqerr` in particular is rendered much harder to
> read than necessary by poor choice of primitives (as above).

Sure, I can introduce smart constructors to make it more readable.

> The definition of `cnn`, much as `forward` before it, uses the new
> 'syntactic' (Agda-/meta-level) representations of E-/object-level
> `let` binders... but it might be useful to offer a side-by-side
> comparison with the deBruijn version, if only for the reader to
> appreciate all the work that has gone into 4.3?

I thought about it, but decided against it as I thought it would drive
too much attention to the syntactic wrappers.  However, after reading
through the reviews, I actually think that emphasising this might be
a good idea.  I am happy to have a go at this.

> p.15--19 AD for E
> 
> Much of this material seems standard by way of describing
> (reverse-mode) AD in terms of straight-line programs defining
> successive partial derivatives in terms of their primals, until the
> mechanics of implementing this on top of `E` is described.
> 
> Surprisingly, on p.17 and the definition of the various auxiliary
> functions as well as the main definition of $\nabla$ (ll.800--825), no
> mention is made of:
> 
> * l.792 that the 'seed' of reverse-mode AD is given by the $\delta$
>   parameter IIUC; so tell the reader!

We have to be really careful with terminology here.  The seed in terms
of reverse mode AD is given by `s` (stands for `s`eed).  Now, $\delta$
is the state of the environment (gradient, -isch) that we are updating.  We have
to call `∇` with some initial value (zero for all positions).  I am happy
to rewrite this sentence.

> * l.823 the use of the HOAS syntax for `let` in the definition of the
>   derivative of `logistic`, so here we do see the full horror of the
>   de Bruijn calculation

Sure, I can mention this.

> * nor anywhere the idea of 'gradient' per se (nor what its meta- nor
>   object- level type might be)

Sufe, I am happy to explain what that is.

> p.19 Section 5.2
> 
> No explicit link is made between the axiomatisation of units for
> addition and multiplication, and the corresponding semantic
> equivalence between 'smart' constructors for addition and
> multiplication and their vanilla counterparts?

Happy to make this explicit.

> p.20
> 
> There is some discussion of how to optimise (or rather: how to avoid
> pessimising) sharing in `let` expressions and their derivatives. The
> paper does not have a lot to say here, but NB citation [34] contains
> extensive discussion of this problem.

Thanks, I can make the link more explicit in the discussion.

> p.21
> 
> Codegen via `String`, IIUC. Perhaps that is the best one can do, but
> it seems as though an Agda-internal DSL for 'Futhark' (scare quotes to
> distinguish for actual Futhark) could be distinguished from `E`
> (esp. wrt higher-rank computations?), so that the questions about
> normalisation of `E` programs into 'Futhark' could be pursued as
> future work?

Absolutely, see our original rebuttal response to this.  In the ideal
world we would introduce embedding of Futhark in Agda, semantics for it
and prove that our normalisation/compilation is semantics-preserving.
This is very obvious future work, but it is not on a critical path
in the following sense: Futhark's syntax is somewhat similar to Agda
and the entire normalisation/extraction takes about 50 lines of code.
Therefore, given that we can compile and run extracted programs, the
chances of mistakes in the translation process is relatively small.
Once again, by no means we claim that this shouldn't be done, we
simply didn't have enough resources to do this in this iteration.

> p.22--23 Section 6
> "One of the goals of this work"
> It would have been helpful to have posted this research question
> upfront in such a way that it would not get lost during the main
> thrust of the paper, concerning the Agda/E/Futhark pipeline.

Sure, happy to stress it more.

> What can we really learn from the performance study? It seems to be
> the conclusion that this pipeline approach is morally good enough,
> without really being clear about where future sources of
> improvement/optimisation might come from. This reader is happily
> convinced by the prehistory of FP that while implementations may be
> slow(er) now, they will improve, but lurking behind the scenes is
> a different problem: namely that of scale wrt number of dimensions/size
> of rank of the tensor calculations involved. In principle, with a
> rank-polymorphic and rank-safe language, such considerations are
> somehow outside the discussion, but when comparing with TensorFlow or
> PyTorch, this is where the sharp distinctions will be observed? I'm
> sure it's out-of-scope for this paper, but some words in this direction
> might have been... helpful to the authors' cause.

Sure, happy to expand on this.

> There is also a separate, unanalysed, question: why extract to Futhark?
> How hard would it be to have a different pipeline/backend, targeting
> PyTorch or TensorFlow directly? What would be involved? how would the
> claimed advantages deriving from rank-polymorphic, rank-safe languages
> in Agda give rise to 'better' back-end computations on other
> platforms? It would be very interesting to know!

Yes, we should say a bit more than what we said in lines 1114-1115,
happy to expand on this.

> p.23--25 Section 7
> Mention Abbott et al.
> Mention Allais et al.

Happy to do this!

> Nevertheless, despite all of the above detailed criticisms, I think
> they can all (mostly!?) be dealt with in such a way as to make a good
> paper much better!

Thanks for all your effort, this is enormously helpful.

### 2025-05-07 rebuttal/response-to-E.md
syntax -> semantics

@@
 This is very obvious future work, but it is not on a critical path
-in the following sense: Futhark's syntax is somewhat similar to Agda
+in the following sense: Futhark's semantics is somewhat similar to Agda
 and the entire normalisation/extraction takes about 50 lines of code.

### 2025-05-27 paper/submitted/intro.tex
Hacky latexdiff setup.

--- 2025-02-27 paper/intro.tex
@@
 guaranteeing the absence of out-of-bound indexing.
-This is a very common source of errors that can be incredibly difficult
+This is a very common source of errors that can be difficult
 to find.   Secondly, we have to compute derivatives of the given tensor expressions,
@@
 \begin{enumerate}
-  \item a rank-polymorphic array theory in Agda;
-  \item an implementation of the CNN from~\cite{cnn-array} in Agda;
+  \item a rank-polymorphic array theory and implementation of
+        the CNN from~\cite{cnn-array} in Agda;
   \item an embedded DSL in Agda which supports AD (reverse mode);

### 2025-05-27 paper/submitted/performance.tex
Hacky latexdiff setup.

--- 2025-02-26 paper/performance.tex
@@
 Our PyTorch version results in extremely high runtimes of 19 and 61
-seconds, and we observe low GPU utilization. While this is likely to
-be a problem with a system configuration, we report these numbers here
-as an indication that performance of machine learning systems is
+seconds, and we observe low GPU utilization. While this likely can be
+improved by adjusting system configuration, we report these numbers here
+as an indication that performance of box-standard machine learning systems is
 fragile.
@@
 \caption{Training time for various training set sizes, comparing
-  Futhark, Tensorflow, and PyTorch, running on an A100 GPU. The final column
+  Futhark and Tensorflow, running on an A100 GPU. The final column
   shows the speedup of TensorFlow compared to Futhark. We use a batch

### 2025-05-28 paper/relatedwork.tex
Address some reviewer A comments.

@@
 does not give a formal proof that AD implementation is correct (it is
-not clear how to formulate correcntess criteria for it), the entire
-implementatuion of AD is about 20 lines of code, so it is easy to
+not clear how to formulate correctness criteria for it), the entire
+implementation of AD is about 20 lines of code, so it is easy to
 understand and modify.
@@
 through Agda's foreign function interface, which posed various
-unresolved implementation challenges. Even though our for support rank
+unresolved implementation challenges. Even though our support for rank
 polymorphism results in having even more complicated types, our code
@@
 
-Verifying index operations is a classic application of dependent
-types, as seen in for example Dependent
-ML~\cite{10.1145/292540.292560}, although most work assumes static
-ranks. Our target language, Futhark, supports size-dependent
-types~\cite{10.1145/3609024.3609412}, but this information is used
-only to ensure shape conformance (e.g., that the operands to a matrix
-multiplication have the same size), not to ensure the correctness of
-array indexing. Single-Assignment C~\cite{sac2} (SaC) supports rank
-polymorphism, including dynamic ranks, but does not guarantee that
-absence of indexing errors.
+Verifying index operations is a classic application of dependent types, as seen
+in for example Dependent ML~\cite{10.1145/292540.292560}, although most work
+assumes static ranks. Our target language, Futhark, supports size-dependent
+types~\cite{10.1145/3609024.3609412}, but this information is used only to
+ensure shape conformance (e.g., that the operands to a matrix multiplication
+have the same size), and does not guarantee that array indexing is in-bounds.
+Single-Assignment C~\cite{sac2} (SaC) supports rank polymorphism, including
+dynamic ranks, but does not guarantee that absence of indexing errors.
 
-Gibbons showed how to express rank polymorphism in Haskell, through
-the use of \emph{Naperian functors}~\cite{10.1145/2976022.2976023}.
-This form of rank polymorphism is more limited than ours; in
-particular, it imposes static ranks. Further, Gibbons' work focused
-solely on the embedding in Haskell, while our system also supports
-code extraction and demonstrates good real-world performance.
-Similarly, Remora~\cite{rank-poly} is a dependently typed
-rank-polymorphic language that also supports only static ranks.
+Gibbons showed how to express rank polymorphism in Haskell, through the use of
+\emph{Naperian functors}~\cite{10.1145/2976022.2976023}. This form of rank
+polymorphism is more limited than ours; in particular, it imposes static ranks.
+This is also the case for libraries that provide rank polymorphic array
+operations, such as Accelerate~\cite{10.1145/1926354.1926358} and
+Repa~\cite{10.1145/1932681.1863582}. Further, Gibbons' work focused solely on
+the embedding in Haskell, while our system also supports code extraction and
+demonstrates good real-world performance. Similarly, Remora~\cite{rank-poly} is
+a dependently typed rank-polymorphic language that also supports only static
+ranks.

### 2025-05-30 paper/intro.tex
Pass throught the paper.
Reviewer D feedback.

@@
 that is competitive with TensorFlow~\cite{ad-tf} and PyTorch~\cite{ad-pytorch}.
-Focussing on correctness, we propose
+Focusing on correctness, we propose
 a theory of rank-polymorphic arrays~\cite{rank-poly} in Agda~\cite{agda-2-6-3}.

### 2025-06-04 paper/relatedwork.tex
More related work.

@@
 how to add AD to a programming language of choice. Examples of these include
-Futhark~\cite{futhark/sc22ad}, Haskell~\cite{ad-haskell}, and
-Jax~\cite{ad-jax,radul2023you}. Furthermore, a number of machine learning
+Futhark~\cite{futhark/sc22ad} and Haskell~\cite{ad-haskell}.
+%
+Furthermore, a number of machine learning
 frameworks that incorporate AD have been proposed in recent years: TensorFlow~\cite{ad-tf},
 PyTorch~\cite{ad-pytorch}, MXNet~\cite{ad-mxnet} and many more.
+%
+Most of these approaches are largely similar to ours (although naturally more
+complete), as they involve a syntactically driven transformation of the program.
+
 While in particular the dedicated frameworks tend to find widespread 
@@
 
-Another line of work studies high-level principles of AD using
-category theory~\cite{ad-theor1, ad-theor2, ad-theor3}.
-While this indeed comes with great correctness guarantees due to
-some naturality principles, it is not always clear how to implement
-this in a way that leads to efficiently executable specifications.  Also, the
-entire treatment of index-safe tensors is unclear.
+The AD transformations in Dex~\cite{10.1145/3473593} and Jax~\cite{ad-jax} are
+in this respect interesting, as they are based on first \emph{linearising} the
+function of interest in a given point (essentially a forward mode
+transformation), then \emph{transposing} the resulting linear map to obtain a
+reverse mode linear map~\cite{radul2023you}. This approach might make be easier
+to implement in a fully verified manner, as linearisation and transposition have
+well-defined properties and could be verified independently.
+
+Another line of work studies high-level principles of AD using category
+theory~\cite{ad-theor1, ad-theor2, ad-theor3}. While this indeed comes with
+great correctness guarantees due to some naturality principles, it is not always
+clear how to implement this in a way that leads to efficiently executable
+specifications, although some efforts are underway~\cite{10.1145/3632878}. Also,
+the entire treatment of index-safe tensors is unclear.
 

### 2025-06-04 paper/intro.tex
Rewrite introduction.

@@
 
-The year is 2025, and we still have to make a formidable choice between
-correctness and performance for all the programming projects that we start.
-Low level programming languages such as C or Fortran make it possible to leverage
-intricate hardware features for the price of poor analysability and
-correctness guarantees.  Dependently-typed systems such as Lean or Agda make it possible
-to describe arbitrary invariants within the given program, yet they can
-rarely generate high-performance code.
+The year is 2025, and programmers still face a trade-off between correctness and
+performance. Low-level languages like C or Fortran allow careful control over
+hardware, but offer few guarantees about safety or program structure.
+Dependently-typed languages like Agda or Lean, on the other hand, make it
+possible to encode precise invariants, but typically lack the infrastructure for
+high-performance computation.
 
-The idealist inside of us is exasperated, because there should be a
-perfect solution that caters for both cases. However, the practitioner
-within us proposes a different perspective. Instead of using a single
-language, we may use two languages in cooperation. Specifically, we
-envision an alliance between a proof assistant and some
-high-performance language of our choice, where we extract code from
-the latter to the former.
+Rather than chasing a mythical language that promises both complete correctness
+and high performance, we explore a practical compromise: leveraging a
+dependently-typed proof assistant for the parts of a scientific programming
+pipeline where full verification is most crucial, and integrating it with a
+high-performance backend for execution. Specifically, we investigate how Agda
+can be used not only for specification and reasoning, but also for
+transformation, optimisation, and code generation in a realistic machine
+learning workload.
 
-To explore this idea, we investigate the concrete problem of automatic differentiation (AD) which
-is often found in machine learning applications.  This is a convenient case study
-as it comes with the following challenges.  From the correctness perspective,
-it is crucially important to track the shapes and ranks of the tensors,
-guaranteeing the absence of out-of-bound indexing.
-This is a very common source of errors that can be difficult
-to find.   Secondly, we have to compute derivatives of the given tensor expressions,
-preserve safe indexing guarantees while we do so, and we have to be able to translate
-the computed expressions into some high-performance language.
-As machine learning applications
-are known to be numerically intensive problems, our performance challenge lies
-in running the program as fast as we can on the chosen hardware architecture.
+To explore this idea, we investigate the concrete problem of automatic
+differentiation (AD) which is often found in machine learning applications. This
+is a convenient case study as it comes with the following challenges. From the
+correctness perspective, it is crucially important to track the shapes and ranks
+of the tensors, guaranteeing the absence of out-of-bound indexing. This is a
+very common source of errors that can be difficult to find. Secondly, we have to
+compute derivatives of the given tensor expressions, preserve safe indexing
+guarantees while we do so, and we have to be able to translate the computed
+expressions into some high-performance language. As machine learning
+applications are known to be numerically intensive problems, our performance
+challenge lies in running the program as fast as we can on the chosen hardware
+architecture.
+
+In an ideal world, all parts would of course come with full formal
+specifications and accompanying proof, but in practice some parts remain an
+entirely separate and nontrivial research effort to verify for functional
+correctness, such as AD. Other parts---such as code generation--can be handled
+using known techniques, but are somewhat time consuming. Despite all this, we
+show that even in those cases where resources preclude full verification, Agda
+can still be used as a practical tool, with full verification employed where it
+is deemed desirable and practical.
 
 We follow~\cite{cnn-array} which demonstrates that it is possible to implement
-one of the canonical convolutional neural network (CNN) in the array
-language SaC~\cite{sac1, sac2}, obtaining good sequential and parallel performance
-that is competitive with TensorFlow~\cite{ad-tf} and PyTorch~\cite{ad-pytorch}.
-Focusing on correctness, we propose
-a theory of rank-polymorphic arrays~\cite{rank-poly} in Agda~\cite{agda-2-6-3}.
-Within this framework, we encode the CNN from~\cite{cnn-array} and lift it 
-into an embedded DSL.  We implement AD
-(reverse mode) and domain-specific optimisations for expressions in that DSL.
-Finally, we implement an extraction 
-into Futhark (a functional array language), apply the CNN to the MNIST digit recognition problem,
-and compare performance with TensorFlow.
+one of the canonical convolutional neural network (CNN) in the array language
+SaC~\cite{sac1, sac2}, obtaining good sequential and parallel performance that
+is competitive with TensorFlow~\cite{ad-tf} and PyTorch~\cite{ad-pytorch}.
+Focusing on correctness, we propose a theory of rank-polymorphic
+arrays~\cite{rank-poly} in Agda~\cite{agda-2-6-3}. Within this framework, we
+encode the CNN from~\cite{cnn-array} and lift it into an embedded DSL. We
+implement AD (reverse mode) and domain-specific optimisations for expressions in
+that DSL. Finally, we implement an extraction into Futhark (a functional array
+language), apply the CNN to the MNIST digit recognition problem, and compare
+performance with TensorFlow.
 
 As a result, we demonstrate an approach where the entire specification,
-optimiser, AD and code generation are available to us within a proof
-assistant of our choice.  We can prove facts about all the
-stages of the pipeline and easily adjust them to our liking.
-We argue that such a liberating
-approach is feasible in practice, at least for the times of dialectic
-of correctness and performance.
+optimiser, AD and code generation are available to us within a proof assistant
+of our choice. We can prove facts about all the stages of the pipeline and
+easily adjust them to our liking. We also demonstrate that some components can
+be left only partially verified, where the effort of verification is judged
+disproportionate to the benefit. We argue that such a liberating approach is
+feasible in practice, at least for the times of dialectic of correctness and
+performance.
 

### 2025-06-11 paper/relatedwork.tex
Reference verification paper.

@@
 
-While in particular the dedicated frameworks tend to find widespread 
-acceptance by practitioners, both, correctness and performance leave
-two open questions: (i) how is it possible to
-ensure that the AD algorithm is implemented correctly?
-(ii) if the
-language does not perform as expected, what are the
-options to solve this?  Unfortunately, in many cases the answer to
-both questions is unsatisfying.  Most languages do not
-come with formal correctness guarantees, so one has to trust the
-implementers of these tools.  One can run tests as well to gain trust 
-in the implementation but that is far from a 
-formal guarantee.  If one relies
-on the AD provided by a chosen language/framework, and the generated code does not
-perform well, one has to modify the language/framework, as these solutions
-are tightly integrated with the tools. The problem here is that most of of these tools
-have very large and sophisticated implementations typically comprising
-of hundreds of thousands of lines of code.  Note that while our work
-does not give a formal proof that AD implementation is correct (it is
-not clear how to formulate correctness criteria for it), the entire
-implementation of AD is about 20 lines of code, so it is easy to
-understand and modify.
+While in particular the dedicated frameworks tend to find widespread acceptance
+by practitioners, both, correctness and performance leave two open questions:
+(i) how is it possible to ensure that the AD algorithm is implemented correctly?
+(ii) if the language does not perform as expected, what are the options to solve
+this? Unfortunately, in many cases the answer to both questions is unsatisfying.
+Most languages do not come with formal correctness guarantees, so one has to
+trust the implementers of these tools. One can run tests as well to gain trust
+in the implementation but that is far from a formal guarantee. If one relies on
+the AD provided by a chosen language/framework, and the generated code does not
+perform well, one has to modify the language/framework, as these solutions are
+tightly integrated with the tools. The problem here is that most of of these
+tools have very large and sophisticated implementations typically comprising of
+hundreds of thousands of lines of code. Note that while our work does not give a
+formal proof that AD implementation is correct, the entire implementation of AD
+is about 20 lines of code, so it is easy to understand and modify. Formulating
+correctness criteria for AD algorithms is somewhat involved, but has been done
+for implementations where the AD algorithm monitors the program when it runs and
+records a trace of its actions~\cite{de2023verifying}, while our approach is
+based on static program transformation---this is in fact the only viable
+approach when also doing program extraction.
 

### 2025-06-12 paper/intro.tex
Emphasize this.

@@
 feasible in practice, at least for the times of dialectic of correctness and
-performance.
+performance. The overall goals is to demonstrate how to formulate a
+computationally intensive problem in a proof assistant, transform it with
+verification of key safety properties, and then pass it on to the other system
+that can run the algorithm efficiently.
 
@@
   \item an extraction mechanism for generating Futhark code from the DSL; and
-  \item an experimental evaluation of the generated codes.
+  \item an experimental evaluation of the generated code.
 \end{enumerate}

### 2025-06-12 paper/performance.tex
Expand.

@@
 
+Another option is of course to generate TensorFlow, or some other established
+high performance language, directly. This is in principle viable, as our overall
+approach is not particularly wedded to Futhark. Extraction does become more
+challenging, as \AF{E} is a \emph{pointed} higher-order
+language~\cite{10.1145/3473593}, meaning programs consists of scalar expressions
+nested inside parallel constructs that explicitly index arrays, while the
+TensorFlow programming model uses first-order bulk parallel constructs that
+implicitly operate on entire arrays at a time.
+
 \begin{table}
